---
title: "MasterControl QAL Analysis - EDA"
author: "Rob Sheehy"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Business Context

MasterControl wants to improve their Mx (Manufacturing) product performance. Right now Mx converts at 12.7% while Qx (Quality) converts at 19.7%. The goal is to figure out which industries, company sizes, and buyer personas actually convert for Mx so they can focus their outbound efforts better.

My focus for this individual EDA: [Pick one - Mx deep dive / Title patterns / Missing data / Segment identification]

---

# Chunk 1: Load and Basic Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(knitr)

# Load data
df <- read_csv("QAL Performance for MSBA.csv", show_col_types = FALSE)

# Quick check
cat("Loaded", nrow(df), "leads from", 
    min(df$`qal cohort date`), "to", max(df$`qal cohort date`), "\n")
```

**Notes - Initial Load**

Data looks clean on import. Need to check for:
- Missing values (I know there's ~40% missing on some fields)
- How the outcome variable is coded
- Whether dates parsed correctly

---

# Chunk 2: Target Variable Check

```{r target-check}
# What outcomes do we have?
outcome_counts <- df %>%
  count(`next_stage__c`, name = "n") %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))

kable(outcome_counts, caption = "Lead Outcomes Distribution")

# Calculate baseline SQL conversion
baseline <- df %>%
  summarise(
    total = n(),
    sql = sum(`next_stage__c` == "SQL", na.rm = TRUE),
    rate = round(sql / total * 100, 2)
  )

cat("\nBaseline SQL conversion rate:", baseline$rate, "%\n")
```

**Notes - Target Variable**

Key findings:
- 71.8% of leads are Recycled - that's the majority outcome
- Only 16.2% make it to SQL 
- Just 42 Won deals (0.2%)

Baseline SQL conversion rate: 16.16%

This tells me most leads don't progress, which aligns with MasterControl's "change is hard" objection. The 12.7% Mx rate is about 3.5 percentage points below baseline - a 20%+ relative gap. Need to find where Mx performs at or above this 16% mark.

Also 2.9% have missing outcomes - will exclude those since I can't classify them.

---

# Chunk 3: Missing Data Assessment

```{r missing-data}
# Calculate missingness
missing_summary <- df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(
    total = nrow(df),
    pct_missing = round(100 * n_missing / total, 1)
  ) %>%
  filter(n_missing > 0) %>%
  arrange(desc(pct_missing))

kable(missing_summary, caption = "Fields with Missing Data")

# Visualize
missing_summary %>%
  ggplot(aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Missing Data by Field",
       x = "Field", y = "% Missing") +
  geom_text(aes(label = paste0(pct_missing, "%")), hjust = -0.1, size = 3)
```

**Notes - Missing Data Challenge**

The big issue: 40.9% missing for manufacturing model and site function. That's almost half the dataset - AI web scraping (Tably AI) failed on these fields.

Also 38.1% missing job titles.

**What this means:**

Any segmentation by manufacturing model or site function only covers 59% of the data. Need to decide: exclude the missing 41%, create "Unknown" category, or try to impute?

For now keeping them but will flag when missingness is high. Need to check if missing data is random or patterned - tells me if it's safe to exclude.

Good news: industry (0.3% missing), tier (0.3%), and territory (0.8%) are nearly complete. Can do basic segmentation with those.

**Bottom line:** This 40% missing will limit my manufacturing model analysis throughout the EDA.

---

# Chunk 4: Product Performance - The Core Problem

```{r mx-qx-comparison}
# Classify leads into Mx, Qx, Other
df_classified <- df %>%
  mutate(
    product_group = case_when(
      solution %in% c("Mx", "Manufacturing") ~ "Mx",
      solution %in% c("Qx", "Quality") ~ "Qx",
      TRUE ~ "Other"
    )
  )

# Calculate conversion by product
product_perf <- df_classified %>%
  filter(product_group %in% c("Mx", "Qx")) %>%
  group_by(product_group) %>%
  summarise(
    total = n(),
    sql = sum(`next_stage__c` == "SQL", na.rm = TRUE),
    rate = round(sql / total * 100, 2),
    recycled = sum(`next_stage__c` == "Recycled", na.rm = TRUE),
    recycled_pct = round(100 * recycled / total, 1)
  )

kable(product_perf, caption = "Mx vs Qx Performance")

# Plot it
product_perf %>%
  ggplot(aes(x = product_group, y = rate, fill = product_group)) +
  geom_col() +
  geom_hline(yintercept = baseline$rate, linetype = "dashed", color = "red") +
  geom_text(aes(label = paste0(rate, "%")), vjust = -0.5) +
  labs(title = "SQL Conversion: Mx vs Qx",
       subtitle = "Red line = Overall baseline (18%)",
       x = NULL, y = "SQL Conversion Rate (%)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Notes - Confirming the Mx Problem**

Numbers confirm the gap:
- Mx: 11.12% conversion ← Problem
- Qx: 17.38% conversion ← Above baseline
- Gap: 6.3 percentage points

Also notice Mx has higher recycled rate (76.7%) vs Qx (70.7%). So Mx leads are more likely to drop off without progressing.

The chart makes it clear - Qx is basically at baseline (18%), but Mx is significantly below at 11.12%.

**Key question:** WHERE does Mx actually perform well? There must be some segments where it works. Need to find those and figure out what's common.

Even getting Mx to 14-15% would be a meaningful improvement - that's a 25-35% relative increase in conversion. The gap looks fixable if I can identify the right target segments.
---

# [YOUR FOCUS AREA STARTS HERE]

## Option A: If focusing on Mx Deep Dive

# Chunk 5: Mx Performance by Industry

```{r mx-by-industry}
# Look at Mx performance across industries
mx_industry <- df_classified %>%
  filter(product_group == "Mx", !is.na(`acct target industry`)) %>%
  group_by(`acct target industry`) %>%
  summarise(
    total = n(),
    sql = sum(`next_stage__c` == "SQL", na.rm = TRUE),
    rate = round(100 * sql / total, 1)
  ) %>%
  filter(total >= 20) %>%  # Need minimum sample
  arrange(desc(rate))

kable(head(mx_industry, 15), 
      caption = "Top 15 Industries for Mx Conversion (min 20 leads)")

# Visualize top 10
mx_industry %>%
  head(10) %>%
  ggplot(aes(x = reorder(`acct target industry`, rate), y = rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  geom_hline(yintercept = 12.7, linetype = "dashed", color = "red") +
  labs(title = "Where Mx Actually Works (Top 10 Industries)",
       subtitle = "Red line = Mx baseline (12.7%)",
       x = NULL, y = "SQL Conversion Rate (%)")
```

**Notes - Mx Winners and Losers**

Looking at where Mx actually performs:
- Blood & Biologics: 15.5% - best performer, above the 12.7% baseline
- Medical Device: 13.4% - also above baseline
- Pharma & BioTech: 10.4% - below baseline despite being the largest volume (1,758 leads)
- Non-Life Science: 10.4% - also underperforming

**Interesting pattern:** The top two performers (Blood & Biologics, Medical Device) are both in highly regulated manufacturing sectors. Makes sense - they have complex manufacturing operations that need real MES systems.

**Volume consideration:** Blood & Biologics looks great at 15.5% but only has 84 leads total. That's not enough volume for a scalable targeting strategy - remember MasterControl's "better 25% with 1,000 than 50% with 50" principle.

Medical Device is more promising: 861 leads with 13.4% conversion. That's both above baseline AND sufficient volume.

Pharma is disappointing - it's the biggest segment (1,758 leads) but only converts at 10.4%. Either Mx isn't a good fit for pharma companies, or we're targeting the wrong personas within pharma.

**Key takeaway:** Mx isn't universally bad - it works in Medical Device. Just needs better targeting to avoid low-converting segments like Pharma and Non-Life Science.
---

# Chunk 6: Mx Performance by Company Size

```{r mx-by-tier}
# First create df_classified (in case it's not in environment)
df_classified <- df %>%
  mutate(
    product_group = case_when(
      solution %in% c("Mx", "Manufacturing") ~ "Mx",
      solution %in% c("Qx", "Quality") ~ "Qx",
      TRUE ~ "Other"
    )
  )

# Now analyze Mx by company tier
mx_tier <- df_classified %>%
  filter(product_group == "Mx",
         !is.na(`acct tier rollup`),
         `acct tier rollup` != "Other") %>%
  group_by(`acct tier rollup`) %>%
  summarise(
    total = n(),
    sql = sum(`next_stage__c` == "SQL", na.rm = TRUE),
    rate = round(100 * sql / total, 1)
  ) %>%
  arrange(desc(rate))

kable(mx_tier, caption = "Mx Conversion by Company Size")

# Plot
mx_tier %>%
  ggplot(aes(x = reorder(`acct tier rollup`, rate), y = rate)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 12.7, linetype = "dashed", color = "red") +
  geom_text(aes(label = paste0(rate, "%")), vjust = -0.5) +
  labs(title = "Mx Conversion by Company Size",
       x = "Company Tier", y = "SQL Conversion Rate (%)")
```

**Notes - Company Size Impact**

Surprising finding: **Small companies convert best** for Mx at 15.6% - above the 12.7% baseline.
- Small: 15.6% (846 leads)
- Medium: 10.9% (2,281 leads)
- Large: 5.5% (542 leads)

This is the opposite of what I expected. Figured bigger manufacturers would have more complex operations needing MES software.

**Possible explanations:**
- Small companies might be more agile, easier to sell to (fewer decision-makers?)
- Large companies might already have MES systems in place - harder to displace
- Medium/Large might have more complex buying processes that kill deals

**Volume consideration:** 
Small companies have both the best conversion (15.6%) AND decent volume (846 leads). Medium has the most volume (2,281) but converts poorly (10.9%).

Using MasterControl's "25% with 1,000 vs 50% with 50" principle: Small companies look like the sweet spot - above-baseline conversion with sufficient volume.

**This changes my thinking:** I assumed we'd want to target Large manufacturers for Mx, but the data says focus on Small companies instead. Need to understand why Large underperforms so badly (5.5% is less than half the baseline).

# Chunk 5: Title Data Quality Check

```{r title-quality}
title_summary <- df %>%
  summarise(
    total = n(),
    with_title = sum(!is.na(`contact/lead title`)),
    missing = sum(is.na(`contact/lead title`)),
    pct_missing = round(100 * missing / total, 1),
    unique_titles = n_distinct(`contact/lead title`, na.rm = TRUE)
  )

cat("Title Summary:\n")
cat("  Total leads:", comma(title_summary$total), "\n")
cat("  With titles:", comma(title_summary$with_title), "\n")
cat("  Missing:", comma(title_summary$missing), 
    paste0("(", title_summary$pct_missing, "%)"), "\n")
cat("  Unique titles:", comma(title_summary$unique_titles), "\n")
```

**Notes - Title Challenge**

5,790 unique title variations. That's the problem - no standardization.

38.1% are missing entirely (6,408 records). For the 10,407 that do have titles, need to extract patterns:
- Seniority (C-Suite, VP, Director, Manager, IC)
- Function (Quality, Operations, Engineering, Regulatory)

This is going to require keyword matching. Can't manually categorize 5,790 variations.

The missing 38% will limit title-based segmentation - any analysis by job title only covers 62% of leads. Similar issue to the manufacturing model missing data.

**Next step:** Look at most common titles to identify keyword patterns I can use for classification.iuk
---

# Chunk 6: Most Common Titles

```{r common-titles}
# Top 30 titles
top_titles <- df %>%
  filter(!is.na(`contact/lead title`)) %>%
  count(`contact/lead title`, sort = TRUE) %>%
  head(30)

kable(top_titles, caption = "Top 30 Job Titles by Frequency")

# Check for keyword patterns
keywords <- c("quality", "manager", "director", "engineer", "vp", "president", 
              "operations", "manufacturing", "regulatory", "senior")

# Count how often each keyword appears
keyword_counts <- tibble(
  keyword = keywords,
  count = map_dbl(keywords, ~sum(str_detect(
    tolower(df$`contact/lead title`[!is.na(df$`contact/lead title`)]), .x
  )))
) %>%
  arrange(desc(count))

kable(keyword_counts, caption = "Keyword Frequency in Titles")
```

**Notes - Title Patterns**

Clear patterns emerging from the top 30 titles and keyword analysis:

**Most common titles are Quality-focused:**
- Quality Manager (252)
- Manager, Quality (124)
- Quality Assurance Manager (98)
- QA Manager (72)

Makes sense given MasterControl's customer base, but this is a problem for Mx - these are Quality personas, not Manufacturing personas.

**Keyword frequency shows I can build a classification system:**

**For Seniority:**
- manager: 3,119 occurrences
- director: 1,746
- senior: 1,082
- president: 458
- vp: 89

**For Function:**
- quality: 3,096 (dominant)
- operations: 774
- manufacturing: 477
- regulatory: 460
- engineer: 987

**Classification approach:** Build hierarchical logic:
1. Check for C-Suite keywords (CEO, President, Founder) → C-Suite
2. Check for VP → VP
3. Check for Director → Director  
4. Check for Manager → Manager
5. Check for Senior → Senior IC
6. Everything else → IC or Unknown

Similar approach for function - check for quality, operations, manufacturing, regulatory, engineering keywords.

**Key insight:** Quality titles dominate (3,096 "quality" vs 477 "manufacturing"). This explains the Mx problem - we're reaching Quality personas who care about QMS, not Manufacturing personas who need MES.

---

## Option C: Segment Identification

# Chunk 5: Combined Segmentation (Industry × Size × Product)

```{r segment-analysis}
# Find high-performing combinations
segments <- df_classified %>%
  filter(!is.na(`acct target industry`),
         !is.na(`acct tier rollup`),
         `acct tier rollup` != "Other",
         product_group %in% c("Mx", "Qx")) %>%
  group_by(`acct target industry`, `acct tier rollup`, product_group) %>%
  summarise(
    total = n(),
    sql = sum(`next_stage__c` == "SQL", na.rm = TRUE),
    rate = round(100 * sql / total, 1),
    .groups = "drop"
  ) %>%
  filter(total >= 30) %>%  # Need statistical confidence
  arrange(desc(rate))

# Show top 20
kable(head(segments, 20), 
      caption = "Top 20 Segments (Industry × Size × Product, min 30 leads)")
```

**Notes - High-Performing Segments**

Looking at the top combinations:

**Best overall:** Medical Device + Small + Qx converts at 28.9% (804 leads) - way above baseline.

**Best Mx segments:**
- Medical Device + Small + Mx: 20.3% (192 leads) ← Best Mx performance I've seen
- Pharma + Small + Mx: 19.0% (326 leads) ← Also strong, decent volume

Pattern is clear: **Small companies are where Mx works.** All the top Mx segments are Small. Medium and Large Mx segments are down in the 9-13% range.

**Interesting finding:** Pharma + Small + Mx converts at 19.0%, but earlier I saw overall Pharma + Mx was only 10.4%. The difference? Company size. Small pharma companies work for Mx, medium/large pharma don't.

**Volume consideration:** 
- Medical Device + Small + Mx: 20.3% conversion with 192 leads - above baseline, decent volume
- Pharma + Small + Mx: 19.0% with 326 leads - even better volume

These are actionable segments. Both have sufficient volume (>100 leads) and beat the 12.7% Mx baseline significantly.

**Key takeaway:** Focus Mx outbound on Small companies in Medical Device and Pharma. A
\void Medium/Large companies - they consistently underperform.

;---

# AI Usage Log

I leveraged AI assistance for coding support and exploratory guidance. The business insights and recommendations reflect my analysis of the findings.

---

**END OF EDA**
