---
title: "Analytical Modeling: MasterControl QAL Lead Progression (Mx vs Qx)"
author: "Group-5 {April Greenwood, Amiya Ranjan Sahoo,Caleb Call , Rob Sheehy}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: readable
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
    fig_width: 10
    fig_height: 6
editor_options:
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


# . Introduction

MasterControl provides enterprise software solutions to regulated life sciences organizations through two primary product lines: Qx, focused on quality and compliance management, and Mx, designed to support digital manufacturing operations. While Qx has demonstrated strong adoption and higher lead progression rates, the Mx product line has consistently underperformed relative to the company baseline.

Because manufacturing solutions require specific operational environments and decision-makers, broad targeting strategies may reduce conversion effectiveness. This project aims to investigate whether the current lead targeting approach for Mx is misaligned with the organizations and buyer personas that are most likely to adopt manufacturing-focused solutions.

By analyzing Qualified Account Lead (QAL) data across industries, manufacturing models, site functions, company tiers, geographic territories, and job titles, this study seeks to identify the structural factors that influence lead progression and to determine how the optimal customer profile for Mx differs from that of Qx.


# . Dataset and Target Variable

The dataset used in this analysis contains 16,644 Qualified Account Lead (QAL) records spanning the period 2024–2025. Each record represents a lead interaction associated with a specific account and product solution.

The dataset contains account-level, contact-level, and campaign-related attributes that describe the context of each lead interaction.

Key variables include:

## . Product Information

	•	solution_rollup – Product solution associated with the lead (Mx, Qx, Ax)
	•	next_stage__c – Sales stage outcome

## . Account Characteristics

	•	acct_target_industry
	•	acct_manufacturing_model
	•	acct_primary_site_function
	•	acct_tier_rollup
	•	acct_territory_rollup

## . Contact Information

	•	contact_lead_title
	•	contact_lead_id

## . Campaign and Lead Context

	•	last_tactic_campaign_channel
	•	priority

A binary target variable, progressed, is constructed from the sales stage field.

Lead progression is defined as advancement to SQL, SQO, or Won, representing movement to advanced stages of the sales funnel.

• progressed = 1  → SQL / SQO / Won

• progressed = 0  → all other outcomes


This formulation enables the problem to be framed as a binary classification task for predictive modeling.


# . Analytical Objective

The primary objective of this project is to identify the account characteristics and buyer personas that most strongly influence lead progression for the Mx product line.

Specifically, the analysis seeks to:

	•	Determine which industries and manufacturing models are most suitable for Mx
	•	Identify which site functions indicate active manufacturing environments
	•	Evaluate how company size and geographic territory influence progression
	•	Determine which job titles and functional roles represent true Mx decision-makers
	•	Compare these findings with the progression patterns observed for Qx

Understanding these relationships will help establish a data-driven Ideal Customer Profile (ICP) for Mx and clarify how Mx targeting should differ from Qx targeting.



# . Modeling Approach

This project integrates descriptive, predictive, and prescriptive analytics to translate observed lead progression patterns into actionable insights.

Descriptive Analytics

The descriptive phase examines progression behavior across key business dimensions such as industry, manufacturing model, site function, company tier, geographic region, and job title. This stage identifies high-performing and underperforming segments within the dataset.

Predictive Analytics

Predictive modeling is used to estimate the probability that a qualified lead will progress beyond the initial qualification stage. By modeling lead progression outcomes, the analysis identifies the variables and feature combinations that most strongly influence success.

Prescriptive Analytics

The final phase translates descriptive and predictive findings into actionable recommendations for sales targeting and lead prioritization. This stage focuses on defining the most promising account and persona combinations for the Mx product line.



# . Predictive Modeling Framework

After completing data preparation and feature engineering, several supervised machine learning models will be evaluated to estimate the probability that a qualified lead progresses beyond the initial qualification stage. The objective of this modeling phase is to identify the account characteristics and buyer attributes that most strongly influence lead progression, while also comparing the predictive performance of multiple algorithms.

The modeling framework includes the following models:

## Model 1: Logistic Regression

A baseline interpretable model that estimates the relationship between predictor variables and the probability of lead progression. Logistic regression provides clear coefficient interpretations, allowing identification of features that increase or decrease progression likelihood.

## Model 2: Elastic-Net Logistic Regression (glmnet)

A regularized extension of logistic regression that combines L1 (Lasso) and L2 (Ridge) penalties. Elastic-Net is particularly useful when working with a large number of engineered predictors, as it performs automatic feature selection and stabilizes model estimates in the presence of correlated variables.

## Model 3: Decision Tree (CART)

A rule-based model that partitions the data into hierarchical segments based on predictor variables. Decision trees are useful for identifying interaction effects and uncovering interpretable decision rules that describe high-conversion account and title segments.

## Model 4: Random Forest

An ensemble learning method that constructs multiple decision trees and aggregates their predictions. Random forests reduce overfitting and capture nonlinear relationships between predictors, while also providing measures of variable importance.

## Model 5: Gradient Boosting / XGBoost

A boosted ensemble model that sequentially builds trees to improve predictive performance. Gradient boosting models are particularly effective for structured tabular datasets and often achieve strong predictive accuracy.

To evaluate model performance, the dataset will be divided into training and testing subsets using a stratified train–test split, ensuring that lead progression outcomes are proportionally represented across both samples.

During training, cross-validation will be used to reduce overfitting and improve model generalization. This approach allows model performance to be evaluated across multiple data partitions and helps ensure stability in predictive results.

Model performance will be assessed using standard classification metrics, including:

	•	Accuracy
	•	Precision
	•	Recall
	•	F1 Score
	•	ROC–AUC

These metrics provide complementary perspectives on model performance and allow comparison across algorithms. The goal is to identify a model that balances predictive accuracy with interpretability while providing insights that can support improved lead targeting and segmentation strategies for the Mx product line.


# . Client Deliverables

The final analysis is designed to produce clear, actionable recommendations addressing the client’s key business questions:

	1.	Which account types should MasterControl target for Mx?(This includes identifying the most promising industries, manufacturing models, and site functions.)
	2.	Which job titles convert best for Mx within those account types?(This includes identifying the seniority levels and functional roles associated with higher progression rates.)
	3.	How does the optimal Mx profile differ from the Qx profile?(This comparison helps clarify where product targeting strategies should diverge.)
	4.	Which accounts and personas represent cross-sell opportunities?(Identifying profiles that demonstrate strong progression potential for both product families.)
	5.	What is the expected impact of improved targeting?(Estimating how refined account and persona targeting could improve Mx progression performance relative to the current baseline.)



# . Business Value of the Analysis

Beyond addressing the specific client deliverables, this analysis provides broader strategic value for MasterControl.

The results can support:

	•	improved sales targeting efficiency
	•	better lead prioritization
	•	clearer definition of the Ideal Customer Profile (ICP) for Mx
	•	improved alignment between marketing, sales, and product strategy
	•	identification of cross-sell opportunities across product lines

By establishing a data-driven understanding of which organizations and decision-makers are most likely to adopt Mx, MasterControl can allocate sales resources more effectively and improve overall progression performance for this strategically important product line.



# . Libraries Loading, Data Preparation and Dataset Inspection

Before beginning the modeling process, the required libraries are loaded to support data manipulation, feature engineering, visualization, model development, and evaluation.

The workflow builds on the tidyverse ecosystem used during the exploratory data analysis phase, while incorporating additional libraries commonly used for predictive modeling and machine learning. These packages support tasks such as feature engineering, categorical encoding, model training, and performance evaluation.

Loading these libraries at the start ensures that the analytical environment is fully prepared for the modeling workflow and promotes reproducibility throughout the analysis.


## . Loading Required Libraries

```{r setup,message=FALSE,warning=FALSE}

# --------------------------------------------------
# Data Import & Manipulation
# --------------------------------------------------
library(readr)        # Reading CSV files
library(dplyr)        # Data manipulation
library(tidyr)        # Data reshaping
library(tibble)       # Modern data frames
library(stringr)      # String processing (useful for title feature engineering)
library(forcats)      # Factor handling
library(data.table)   # Fast file reading for large datasets

# --------------------------------------------------
# Data Cleaning & Exploration
# --------------------------------------------------
library(janitor)      # Clean column names
library(skimr)        # Detailed summary statistics
library(lubridate)    # Date handling

# --------------------------------------------------
# Missing Data Analysis
# --------------------------------------------------
library(naniar)       # Missing data visualization

# --------------------------------------------------
# Visualization
# --------------------------------------------------
library(ggplot2)      # Core visualization package
library(scales)       # Percent formatting
library(ggthemes)     # Professional plot themes
library(gridExtra)    # Arrange multiple plots
library(GGally)       # Pair plots and correlation visuals

# --------------------------------------------------
# Table Presentation
# --------------------------------------------------
library(gt)           # Professional formatted tables

# --------------------------------------------------
# Categorical Association
# --------------------------------------------------
library(rcompanion)   # Cramer's V for categorical correlation

# --------------------------------------------------
# Machine Learning & Modeling
# --------------------------------------------------
library(caret)        # Modeling framework and resampling
library(recipes)      # Feature engineering pipelines
library(glmnet)       # Elastic-Net regularized regression
library(rpart)        # Decision tree (CART)
library(rpart.plot)   # Decision tree visualization
library(randomForest) # Random Forest model
library(xgboost)      # Gradient boosting model
library(pROC)         # ROC curves and AUC evaluation

# --------------------------------------------------
# Global Settings
# --------------------------------------------------
theme_set(theme_minimal())   # Set default visualization theme
set.seed(123)                # Ensure reproducibility

```


## . Loading the Dataset

The Qualified Account Lead (QAL) dataset used in this analysis contains lead-level observations describing account attributes, contact information, campaign context, and sales stage outcomes.

Each record represents a qualified account lead (QAL) associated with a specific account, campaign interaction, and product solution. This dataset forms the foundation for the predictive modeling framework developed in the following sections.

```{r, load-data}
# Load the MasterControl QAL dataset
qal_data <- read_csv("Mastercontrol QAL Performance.csv")

qal_data <- qal_data %>% clean_names()
```

## . Initial Dataset Inspection

Before proceeding with modeling preparation, the dataset is inspected to verify that it has been imported correctly and to understand its overall structure. This step provides a quick overview of the dataset’s size, variable names, data types, and general characteristics.

These checks help confirm that the dataset is ready for further preparation and modeling.



```{r, Inspect-dataset}

# ---------------------------
# Dataset Dimensions
# ---------------------------

dim_table <- tibble(
  Metric = c("Number of Rows", "Number of Columns"),
  Value  = c(nrow(qal_data), ncol(qal_data))
)
print(
dim_table %>%
  gt()
)
# ---------------------------
# Column Names
# ---------------------------

column_table <- tibble(
  Column_Name = colnames(qal_data)
)
print(
column_table %>%
  gt()
)
# ---------------------------
# Dataset Preview
# ---------------------------

preview_head <- head(qal_data, 10)
preview_tail <- tail(qal_data, 10)

preview_head %>% gt()
preview_tail %>% gt()

# ---------------------------
# Data Structure
# ---------------------------

structure_table <- tibble(
  Column = names(qal_data),
  Data_Type = sapply(qal_data, class),
  Unique_Values = sapply(qal_data, function(x) length(unique(x)))
)
print(
structure_table %>%
  gt()
)

```



## . Summary Statistics and Data Quality Checks

Before moving to feature engineering and modeling, it is important to evaluate the overall quality of the dataset. This includes examining summary statistics, identifying missing values, and evaluating the number of unique values in each variable.

These checks help identify potential data preparation tasks required before modeling.


```{r, Data-Quality}
# ---------------------------
# Numeric Summary Statistics
# ---------------------------

numeric_data <- qal_data %>%
  select(where(is.numeric))

if(ncol(numeric_data) > 0){

summary_table <- summary(numeric_data) %>%
  as.data.frame() %>%
  rownames_to_column("Statistic")

summary_table %>%
  gt()

} else {

message("No numeric variables found in the dataset.")

}

# ---------------------------
# Detailed Dataset Overview
# ---------------------------

skim(qal_data)

# ---------------------------
# Missing Values Analysis
# ---------------------------

missing_table <- tibble(
  Variable = names(qal_data),
  Missing_Count = colSums(is.na(qal_data)),
  Missing_Percent = round(colSums(is.na(qal_data)) / nrow(qal_data) * 100, 2)
)
print(
missing_table %>%
  arrange(desc(Missing_Percent)) %>%
  gt()
)
# ---------------------------
# Variable Cardinality
# ---------------------------

cardinality_table <- tibble(
  Variable = names(qal_data),
  Unique_Values = sapply(qal_data, function(x) length(unique(x)))
)
print(
cardinality_table %>%
  arrange(desc(Unique_Values)) %>%
  gt()
)
```


## . Data Integrity Checks

Before proceeding to modeling preparation, several additional integrity checks are performed to confirm that the dataset structure is reliable. These checks ensure that column names are standardized, date fields are correctly parsed, and that no unintended duplicate records exist.

The following checks are conducted:

	•	verification of column naming format
	•	validation of date variable type
	•	detection of duplicated rows
	•	confirmation that the primary identifier (qal_id) is unique

These checks help prevent data quality issues that could affect downstream modeling results.


```{r, Data_Integrity_Check}
# --------------------------------------------------
# 1. Column Names
# --------------------------------------------------

column_format_table <- tibble(
  Column_Name = names(qal_data)
)
print(
column_format_table %>%
  gt()
)
# --------------------------------------------------
# 2. Date Variable Type
# --------------------------------------------------

date_type_table <- tibble(
  Variable = "qal_cohort_date",
  Data_Type = class(qal_data$qal_cohort_date)
)
print(
date_type_table %>%
  gt()
)
# --------------------------------------------------
# 3. Duplicate Row Check
# --------------------------------------------------

duplicate_rows <- sum(duplicated(qal_data))

duplicate_table <- tibble(
  Metric = "Duplicate Rows (Full Record)",
  Count = duplicate_rows
)
print(
duplicate_table %>%
  gt()
)
# --------------------------------------------------
# 4. qal_id Uniqueness Check
# --------------------------------------------------

id_check_table <- tibble(
  Metric = c("Total Records", "Unique qal_id", "Duplicate qal_id"),
  Value = c(
    nrow(qal_data),
    n_distinct(qal_data$qal_id),
    nrow(qal_data) - n_distinct(qal_data$qal_id)
  )
)

id_check_table %>%
  gt()

# --------------------------------------------------
# 5. Inspect Duplicate Records (if any)
# --------------------------------------------------

duplicate_records <- qal_data %>%
  filter(duplicated(.))

duplicate_records %>%
  head(10) %>%
  gt()
```



## . Categorical Distribution Overview

Before modeling preparation, it is useful to examine the distribution of major categorical variables to understand class imbalance and rare categories.

```{r,echo=TRUE,results='hide'}
categorical_vars <- qal_data %>%
  select(where(is.character))

for(col in names(categorical_vars)) {
  
  freq_table <- qal_data %>%
    count(.data[[col]], sort = TRUE)
  
  print(
    gt(freq_table) %>%
      tab_header(
        title = paste("Distribution of", col)
      )
  )
}
```

## . Detection of Non-Standard Missing Labels

The dataset was checked for multiple forms of missing information, including standard NA values, non-standard text labels such as:

	•	"Unknown"
	•	"Not Enough Info Found"
	•	"MISSING/BLANK"
	•  Empty cells (””)

 The results confirm that empty cells in the source file were automatically imported into R as NA. Therefore, no additional empty-string values ("") were detected separately in the dataset.

```{r, error=TRUE}
missing_patterns <- bind_rows(
  
  qal_data %>%
    summarise(across(everything(),
                     ~ sum(is.na(.)))) %>%
    mutate(Missing_Type = "NA"),
  
  qal_data %>%
    summarise(across(everything(),
                     ~ sum(. == "Unknown", na.rm = TRUE))) %>%
    mutate(Missing_Type = "Unknown"),
  
  qal_data %>%
    summarise(across(everything(),
                     ~ sum(. == "Not Enough Info Found", na.rm = TRUE))) %>%
    mutate(Missing_Type = "Not Enough Info Found"),
  
  qal_data %>%
    summarise(across(everything(),
                     ~ sum(. == "MISSING/BLANK", na.rm = TRUE))) %>%
    mutate(Missing_Type = "MISSING/BLANK"),
  
  qal_data %>%
    summarise(across(everything(),
                     ~ sum(. == "", na.rm = TRUE))) %>%
    mutate(Missing_Type = 'Empty Cells ("")')
  
) %>%
  relocate(Missing_Type)

missing_patterns %>%
  gt()
```




These inspection steps confirm that the dataset structure, variable types, and missing-value patterns are well understood before proceeding to feature engineering and predictive modeling.



# . Data Preparation for Modeling


After completing the initial data audit, the next step is to prepare the dataset for predictive modeling. This phase focuses on cleaning the data, standardizing missing values, defining the target variable, removing records with undefined outcomes, and creating a structured base dataset that can later be used for feature engineering and model development.

At this stage, the objective is not yet to create engineered features, but rather to ensure that the underlying data is clean, consistent, and suitable for downstream modeling tasks.

The main preparation steps include:

	•	standardizing non-standard missing values
	•	confirming the effect of missing-value conversion
	•	removing records with undefined target outcomes
	•	constructing the binary target variable
	•	creating the modeling base dataset
	•	reviewing the final class distribution


## . Create a Working Copy of the Dataset

Before applying any transformations, a working copy of the cleaned dataset is created. This preserves the original imported dataset and allows all modeling-related preparation steps to be performed in a controlled and reproducible way.

```{r, Create-Working-Copy}

# Creating a working copy for modeling preparation
qal_model_base <- qal_data

```


## . Standardize Non-Standard Missing Labels

The dataset contains several forms of missing information beyond standard NA values. These include text-based placeholders such as “Unknown”, “Not Enough Info Found”, and “MISSING/BLANK”. To ensure consistency during modeling, these values are converted into standard NA values.

This step improves data quality and prevents placeholder labels from being treated as legitimate predictor categories later in the analysis.


```{r, Standardizing-Missing-Labels}

# Converting non-standard missing labels to NA
qal_model_base <- qal_model_base %>%
  mutate(across(
    everything(),
    ~ case_when(
      . %in% c("Unknown",
               "Not Enough Info Found",
               "MISSING/BLANK") ~ NA_character_,
      TRUE ~ as.character(.)
    )
  ))

# Convert qal_cohort_date back to Date format
qal_model_base <- qal_model_base %>%
  mutate(qal_cohort_date = lubridate::mdy(qal_cohort_date))
```




## . Review Missing Values After Standardization

After converting non-standard missing labels into NA, it is useful to reassess the amount of missing data in each variable. This confirms that all missing values are now represented consistently and provides a clean baseline for later feature engineering.

```{r, Missing-Values-After-Standardization}

missing_after_cleaning <- tibble(
  Variable = names(qal_model_base),
  Missing_Count = colSums(is.na(qal_model_base)),
  Missing_Percent = round(colSums(is.na(qal_model_base)) / nrow(qal_model_base) * 100, 2)
)

missing_after_cleaning %>%
  arrange(desc(Missing_Percent)) %>%
  gt()
```



## . Remove Records With Undefined Outcome

The outcome variable for modeling is derived from next_stage_c. Records with missing values in this column do not have a defined sales progression outcome and therefore cannot be used in supervised learning.

These records are removed before creating the target variable.

```{r,Remove-Undefined-Outcome }

# Count rows with undefined target outcome
undefined_outcome_table <- tibble(
  Metric = c("Total Records Before Filtering",
             "Records With Missing next_stage_c",
             "Records Remaining After Filtering"),
  Value = c(
    nrow(qal_model_base),
    sum(is.na(qal_model_base$next_stage_c)),
    sum(!is.na(qal_model_base$next_stage_c))
  )
)

undefined_outcome_table %>%
  gt()

# Remove records with missing next_stage_c
qal_model_base <- qal_model_base %>%
  filter(!is.na(next_stage_c))

```



## . Construct the Target Variable

The modeling objective is to predict whether a lead progressed to a deeper sales stage. A binary target variable named progressed is created from next_stage_c.

Leads that reached SQL, SQO, or Won are classified as progressed (1), while all other observed outcomes are treated as non-progressed (0).

```{r,Construct-Target-Variable}
# Create binary target variable
qal_model_base <- qal_model_base %>%
  mutate(
    progressed = if_else(
      next_stage_c %in% c("SQL", "SQO", "Won"),
      1,
      0
    )
  )

# Review target mapping
target_mapping_table <- qal_model_base %>%
  count(next_stage_c, progressed, sort = TRUE)

target_mapping_table %>%
  gt()
```



## . Review Target Variable Distribution

After the target variable is created, the class distribution should be reviewed to understand the balance between progressed and non-progressed leads. This is important because class imbalance can affect model performance and interpretation.


```{r,Target-Distribution}
target_distribution <- qal_model_base %>%
  count(progressed) %>%
  mutate(
    Percentage = round(n / sum(n) * 100, 2)
  )

target_distribution %>%
  gt()
```



## . Prepare the Modeling Base Dataset

At this stage, the dataset is cleaned and the target variable has been defined. The next step is to prepare the base dataset that will be used in later stages such as feature engineering and train-test splitting.

For now, identifier columns and raw outcome columns are retained for traceability, but they are clearly flagged as variables that will not be used as predictors during model training.

```{r, Modeling-Base-Dataset-Summary}
model_base_summary <- tibble(
  Metric = c("Total Records in Modeling Base Dataset",
             "Total Columns in Modeling Base Dataset"),
  Value = c(nrow(qal_model_base), ncol(qal_model_base))
)

model_base_summary %>%
  gt()

# Preview key columns
qal_model_base %>%
  select(qal_id, contact_lead_id, next_stage_c, progressed, solution_rollup, qal_cohort_date) %>%
  head(10) %>%
  gt()
```


## . Identify Columns for Exclusion From Predictors

Some columns should not be used directly as predictors because they are identifiers or because they contain the raw outcome information used to create the target variable.

These columns will be excluded later during model matrix construction, but they are identified now to ensure a clear modeling workflow.

```{r,Predictor-Execution-List}
predictor_exclusion_table <- tibble(
  Column = c("qal_id", "contact_lead_id", "next_stage_c"),
  Reason = c(
    "Unique record identifier",
    "Unique contact identifier",
    "Raw outcome field used to create the target variable"
  )
)

predictor_exclusion_table %>%
  gt()
```


## . Final Check Before Feature Engineering

Before moving to feature engineering, a final confirmation is made to ensure that the modeling base dataset is complete, the target variable is defined, and missing values have been standardized.

This serves as the final checkpoint before the dataset is transformed into model-ready features.

```{r, Final-Preparation-Check}
final_preparation_check <- tibble(
  Metric = c(
    "Records Available for Modeling",
    "Progressed Class (1)",
    "Non-Progressed Class (0)",
    "Columns with Remaining Missing Values"
  ),
  Value = c(
    nrow(qal_model_base),
    sum(qal_model_base$progressed == 1),
    sum(qal_model_base$progressed == 0),
    sum(colSums(is.na(qal_model_base)) > 0)
  )
)

final_preparation_check %>%
  gt()
```




## . Additional Pre-Modeling Validation Checks

Before proceeding to feature engineering, several additional validation checks are performed to confirm that the modeling dataset is structurally sound. These checks verify the distribution of product types (Mx vs Qx), confirm that the progression target variable behaves as expected across products, ensure that the target variable contains no missing values, validate the date variable format, and review the overall dataset structure.

These checks provide an additional safeguard to ensure that the dataset is ready for feature engineering and subsequent model development.

```{r }
# --------------------------------------------------
# Additional Dataset Validation Checks
# --------------------------------------------------

# 1. Product Distribution (Mx vs Qx)
product_distribution <- qal_model_base %>%
  count(solution_rollup) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

product_distribution %>%
  gt() %>%
  tab_header(
    title = "Distribution of Product Types (Mx vs Qx)"
  )


# 2. Progression Distribution by Product
progression_by_product <- qal_model_base %>%
  count(solution_rollup, progressed) %>%
  group_by(solution_rollup) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2))

progression_by_product %>%
  gt() %>%
  tab_header(
    title = "Progression Distribution by Product"
  )


# 3. Verify No Missing Values in Target Variable
target_missing_check <- tibble(
  Metric = "Missing Values in Target Variable (progressed)",
  Value = sum(is.na(qal_model_base$progressed))
)

target_missing_check %>%
  gt()


# 4. Confirm Date Variable Type
date_type_check <- tibble(
  Variable = "qal_cohort_date",
  Data_Type = class(qal_model_base$qal_cohort_date)
)

date_type_check %>%
  gt()


# 5. Review Dataset Structure
structure_table <- tibble(
  Column = names(qal_model_base),
  Data_Type = sapply(qal_model_base, class),
  Unique_Values = sapply(qal_model_base, function(x) length(unique(x)))
)

structure_table %>%
  gt() %>%
  tab_header(
    title = "Modeling Dataset Structure Overview"
  )

```


# . Feature Engineering

After preparing the modeling base dataset, the next step is to engineer features that improve the predictive power and interpretability of the models. Feature engineering is especially important in this project because several variables are high-cardinality categorical fields, free-text job titles, or date variables that require transformation before they can be used effectively in machine learning models.

The goal of this stage is to convert raw variables into structured, meaningful predictors that better represent account characteristics, buyer roles, and lead context. This step is expected to play a major role in identifying the profiles most likely to progress and in supporting the final business deliverables related to Mx targeting and cross-sell opportunities.

The main feature engineering tasks include:

	•	extracting structured information from job titles
	•	deriving time-based features from cohort dates
	•	grouping rare categories in high-cardinality variables
	•	preserving meaningful missingness where appropriate
	•	creating analysis-ready categorical predictors for modeling
	
	
## . Create a Working Copy for Feature Engineering

Before creating new predictors, a separate working copy of the modeling dataset is created. This preserves the prepared base dataset and allows feature engineering steps to be applied in a controlled and reproducible manner.

```{r}
# --------------------------------------------------
# Create Feature Engineering Working Dataset
# --------------------------------------------------

qal_features <- qal_model_base
```


## . Extract Time-Based Features from qal_cohort_date

The cohort date can capture temporal patterns in lead performance.
Leads from certain quarters or years may convert differently.

Create:

	•	Year
	•	Quarter
	•	Month

```{r}
# --------------------------------------------------
# Time-Based Features
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    cohort_year = lubridate::year(qal_cohort_date),
    cohort_quarter = lubridate::quarter(qal_cohort_date),
    cohort_month = lubridate::month(qal_cohort_date)

  )
```

```{r, year-quarter feature}
qal_features <- qal_features %>%
  mutate(
    cohort_year_quarter = paste0(cohort_year, "-Q", cohort_quarter)
  )
```


## . Preserve Missingness Indicators

Some missing values may carry information (for example missing titles).
Create binary indicators for missing fields.
This allows models to learn whether missing titles behave differently.

```{r}
# --------------------------------------------------
# Missing Value Indicators
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    title_missing = if_else(is.na(contact_lead_title), 1, 0)
  )
```


## .Group Rare Categories in High-Cardinality Variables

Some categorical variables have very small categories that add noise.

We group rare categories into “Other”.

Variables most likely affected:

	•	acct_primary_site_function
	•	acct_manufacturing_model
	•	last_tactic_campaign_channel
	
	This keeps the top frequent categories and groups the rest.
	
```{r}
# --------------------------------------------------
# Group Rare Categories
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    
    acct_primary_site_function =
      forcats::fct_lump_n(acct_primary_site_function, n = 10),
    
    acct_manufacturing_model =
      forcats::fct_lump_n(acct_manufacturing_model, n = 10),
    
    last_tactic_campaign_channel =
      forcats::fct_lump_n(last_tactic_campaign_channel, n = 8)
    
  )
```
	
	
## .Convert Core Predictors to Factor Type

Most variables in your dataset are categorical predictors.
Converting them to factors ensures modeling frameworks treat them correctly.The target variable remains numeric (0/1).


```{r}
# --------------------------------------------------
# Convert Categorical Predictors to Factors
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    acct_primary_site_function = as.factor(acct_primary_site_function),
    acct_manufacturing_model = as.factor(acct_manufacturing_model),
    acct_target_industry = as.factor(acct_target_industry),
    acct_territory_rollup = as.factor(acct_territory_rollup),
    acct_tier_rollup = as.factor(acct_tier_rollup),
    solution = as.factor(solution),
    solution_rollup = as.factor(solution_rollup),
    last_tactic_campaign_channel = as.factor(last_tactic_campaign_channel),
    priority = as.factor(priority),
    cohort_year = as.factor(cohort_year),
    cohort_quarter = as.factor(cohort_quarter),
    cohort_month = as.factor(cohort_month),
    cohort_year_quarter = as.factor(cohort_year_quarter)
  )
```


## .Review the Engineered Dataset

Check the structure and confirm new variables.

```{r}
# --------------------------------------------------
# Review Engineered Dataset
# --------------------------------------------------

feature_structure <- tibble(
  Column = names(qal_features),
  Data_Type = sapply(qal_features, class),
  Unique_Values = sapply(qal_features, function(x) length(unique(x)))
)

feature_structure %>%
  gt() %>%
  tab_header(
    title = "Engineered Dataset Structure"
  )
```


## . Final Check Before Title Feature Engineering

Confirm dataset integrity before handling titles.


```{r}
final_feature_check <- tibble(
  Metric = c(
    "Records",
    "Columns",
    "Progressed Leads",
    "Non-Progressed Leads"
  ),
  Value = c(
    nrow(qal_features),
    ncol(qal_features),
    sum(qal_features$progressed == 1),
    sum(qal_features$progressed == 0)
  )
)

final_feature_check %>%
  gt()
```




## .Title Feature Engineering


### .Clean the Title Text

Before extracting structured signals, the raw title text is standardized so that titles with different capitalization, spacing, or punctuation can be interpreted consistently. This step preserves the original title field while creating a cleaned version for feature engineering.

```{r, Title-Cleaning}
# --------------------------------------------------
# Clean Contact Lead Title
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    title_clean = contact_lead_title,
    title_clean = str_to_lower(title_clean),
    title_clean = iconv(title_clean, from = "", to = "ASCII//TRANSLIT"),
    title_clean = str_replace_all(title_clean, "[^a-z0-9/&+ -]", " "),
    title_clean = str_replace_all(title_clean, "\\s+", " "),
    title_clean = str_trim(title_clean)
  )
```


### .Create Seniority Features

The title field contains strong signals about decision-making authority. Seniority is extracted to identify whether the contact is an executive, vice president, director, manager, or an individual contributor.

To ensure that all titles are included, any title that does not match the main seniority rules is assigned to "Other", while missing titles are explicitly preserved as "Missing".

```{r, Title-Seniority-Feature}

# --------------------------------------------------
# Title Seniority
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    title_seniority = case_when(
      
      # Missing titles
      is.na(contact_lead_title) | title_clean == "" ~ "Missing",

      # Vice Presidents (checked first to avoid being captured as Executive)
      str_detect(title_clean,
                 "\\bvp\\b|vice president|svp|evp|senior vice president|executive vice president|assistant vice president|associate vice president") ~ "VP",

      # C-Suite Executives
      str_detect(title_clean,
                 "\\bchief\\b|\\bceo\\b|\\bcfo\\b|\\bcoo\\b|\\bcio\\b|\\bcto\\b|\\bpresident\\b|chief .* officer|founder|co founder|owner|partner|principal") ~ "Executive",

      # Directors
      str_detect(title_clean,
                 "\\bdirector\\b|associate director|senior director|executive director|global director|site director|\\bhead\\b") ~ "Director",

      # Managers
      str_detect(title_clean,
                 "\\bmanager\\b|general manager|senior manager|assistant manager|group manager|project manager|program manager|supervisor|team lead|\\blead\\b") ~ "Manager",

      # Individual contributors
      str_detect(title_clean,
                 "engineer|scientist|specialist|analyst|associate|coordinator|administrator|technician|consultant|representative|advisor|developer|programmer|chemist|pharmacist|assessor|inspector|officer") ~ "Individual_Contributor",

      # Everything else
      TRUE ~ "Other"
    ),
    
    title_seniority = as.factor(title_seniority)
  )

```




### .Create Functional Role Features

In addition to seniority, titles also indicate the department or business function the contact belongs to. These roles are especially important for distinguishing Mx and Qx buyer profiles.

Each title is assigned to one primary functional category. Missing titles are preserved as "Missing", and any title that does not match the major business patterns is assigned to "Other".

```{r, Title_Functional_Role}
# --------------------------------------------------
# Title Functional Role
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    title_function = case_when(
      
      is.na(contact_lead_title) | title_clean == "" ~ "Missing",
      
      str_detect(title_clean,
                 "quality|\\bqa\\b|\\bqc\\b|qms|compliance|validation|document control|sterility assurance|quality systems|quality control|quality assurance") ~ "Quality_Compliance",
      
      str_detect(title_clean,
                 "regulatory|\\bra\\b|reg affairs|regulatory affairs|qualified person|dscsa") ~ "Regulatory",
      
      str_detect(title_clean,
                 "manufacturing|production|plant|process|operations|operational excellence|continuous improvement|lean manufacturing|industrial") ~ "Manufacturing_Operations",
      
      str_detect(title_clean,
                 "engineer|engineering|scientist|research|development|r d|r&d|product development|process development|automation|laboratory") ~ "Engineering_RD",
      
      str_detect(title_clean,
                 "information technology|\\bit\\b|systems|software|digital|technology|data|infrastructure|computer|programmer|developer|architect") ~ "IT_Systems",
      
      str_detect(title_clean,
                 "sales|business development|commercial|marketing|account manager|account executive|product manager|customer success|go to market") ~ "Commercial_Business",
      
      str_detect(title_clean,
                 "supply chain|procurement|logistics|warehouse|cold chain|purchasing|supplier|distribution") ~ "Supply_Chain",
      
      str_detect(title_clean,
                 "clinical|medical|post market|surveillance|patient|pharmacovigilance") ~ "Clinical_Medical",
      
      str_detect(title_clean,
                 "finance|financial|accountant|accounts payable|human resources|\\bhr\\b|administration|admin|management assistant|tax|payroll") ~ "Finance_Admin",
      
      TRUE ~ "Other"
    ),
    
    title_function = as.factor(title_function)
  )

```



### .Create Keyword Flags

Binary keyword flags preserve important signals that may overlap across titles. This is especially useful when one title contains multiple important concepts such as both seniority and department.


```{r, Title_Keyword_Flags}

# --------------------------------------------------
# Title Keyword Flags
# --------------------------------------------------

qal_features <- qal_features %>%
  mutate(
    has_quality        = if_else(!is.na(title_clean) & str_detect(title_clean, "quality|\\bqa\\b|\\bqc\\b|qms|compliance|validation"), 1, 0),
    
    has_regulatory     = if_else(!is.na(title_clean) & str_detect(title_clean, "regulatory|\\bra\\b|reg affairs|qualified person"), 1, 0),
    
    has_manufacturing  = if_else(!is.na(title_clean) & str_detect(title_clean, "manufacturing|production|plant|process|industrial"), 1, 0),
    
    has_operations     = if_else(!is.na(title_clean) & str_detect(title_clean, "operations|operational excellence|continuous improvement"), 1, 0),
    
    has_engineering    = if_else(!is.na(title_clean) & str_detect(title_clean, "engineer|engineering|scientist|research|development|automation"), 1, 0),
    
    has_it             = if_else(!is.na(title_clean) & str_detect(title_clean, "information technology|\\bit\\b|systems|software|digital|data|technology"), 1, 0),
    
    has_sales          = if_else(!is.na(title_clean) & str_detect(title_clean, "sales|business development|commercial|marketing|account"), 1, 0),
    
    has_supply_chain   = if_else(!is.na(title_clean) & str_detect(title_clean, "supply chain|procurement|logistics|warehouse|supplier|purchasing"), 1, 0),
    
    has_executive      = if_else(!is.na(title_clean) & str_detect(title_clean, "\\bchief\\b|\\bceo\\b|\\bcfo\\b|\\bcoo\\b|\\bcio\\b|\\bcto\\b|founder|owner|partner"), 1, 0),
    
    has_vp             = if_else(!is.na(title_clean) & str_detect(title_clean, "\\bvp\\b|vice president"), 1, 0),
    
    has_director       = if_else(!is.na(title_clean) & str_detect(title_clean, "\\bdirector\\b|\\bhead\\b"), 1, 0),
    
    has_manager        = if_else(!is.na(title_clean) & str_detect(title_clean, "\\bmanager\\b|supervisor|team lead|general manager"), 1, 0)
  )
```



### .Review the Title Features

After engineering the title-based predictors, the new variables should be reviewed to confirm that all titles have been assigned correctly and that no rows have been lost. This review also helps confirm that the "Missing" and "Other" fallback categories are working as intended.

```{r, Review_Title-Based_Features}

# --------------------------------------------------
# Review Title-Based Features
# --------------------------------------------------

title_feature_summary <- tibble(
  Metric = c(
    "Total Records",
    "Missing Raw Titles",
    "Unique Raw Titles",
    "Unique Seniority Categories",
    "Unique Functional Categories"
  ),
  Value = c(
    nrow(qal_features),
    sum(is.na(qal_features$contact_lead_title)),
    n_distinct(qal_features$contact_lead_title, na.rm = TRUE),
    n_distinct(qal_features$title_seniority),
    n_distinct(qal_features$title_function)
  )
)

title_feature_summary %>%
  gt()

# Distribution of seniority
qal_features %>%
  count(title_seniority, sort = TRUE) %>%
  gt() %>%
  tab_header(title = "Distribution of Title Seniority")

# Distribution of function
qal_features %>%
  count(title_function, sort = TRUE) %>%
  gt() %>%
  tab_header(title = "Distribution of Title Function")

```




### . Final Coverage Check

A final validation step confirms that every title has been captured by the rule system. Titles should fall into one of three groups:

	•	matched to a specific category
	•	classified as "Other"
	•	classified as "Missing"

This ensures that every title is included in the engineered dataset.

```{r}

title_coverage_check <- tibble(
  Metric = c(
    "Records with Missing Seniority",
    "Records with Missing Function",
    "Records Assigned to Other Seniority",
    "Records Assigned to Other Function"
  ),
  Value = c(
    sum(is.na(qal_features$title_seniority)),
    sum(is.na(qal_features$title_function)),
    sum(qal_features$title_seniority == "Other"),
    sum(qal_features$title_function == "Other")
  )
)

title_coverage_check %>%
  gt()

```


## .Final Review of the Engineered Dataset

This step validates that all engineered features have been successfully created and that the dataset structure is appropriate for modeling. The review includes checks for dataset size, column structure, variable types, and key feature distributions.

```{r}
# --------------------------------------------------
# Final Review of Engineered Dataset
# --------------------------------------------------

# 1. Dataset size check
dataset_summary <- tibble(
  Metric = c(
    "Total Records",
    "Total Columns",
    "Progressed Leads",
    "Non-Progressed Leads"
  ),
  Value = c(
    nrow(qal_features),
    ncol(qal_features),
    sum(qal_features$progressed == 1),
    sum(qal_features$progressed == 0)
  )
)

dataset_summary %>%
  gt() %>%
  tab_header(title = "Engineered Dataset Summary")


# 2. Dataset structure overview
structure_table <- tibble(
  Column = names(qal_features),
  Data_Type = sapply(qal_features, class),
  Unique_Values = sapply(qal_features, function(x) length(unique(x)))
)

structure_table %>%
  gt() %>%
  tab_header(title = "Engineered Dataset Structure")


# 3. Seniority distribution
qal_features %>%
  count(title_seniority, sort = TRUE) %>%
  gt() %>%
  tab_header(title = "Distribution of Title Seniority")


# 4. Functional role distribution
qal_features %>%
  count(title_function, sort = TRUE) %>%
  gt() %>%
  tab_header(title = "Distribution of Title Functional Roles")


# 5. Product distribution (Mx vs Qx)
qal_features %>%
  count(solution_rollup) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2)) %>%
  gt() %>%
  tab_header(title = "Product Distribution (Mx vs Qx)")


# 6. Progression distribution by product
qal_features %>%
  count(solution_rollup, progressed) %>%
  group_by(solution_rollup) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2)) %>%
  gt() %>%
  tab_header(title = "Lead Progression by Product")
```



# .Encoding and Train/Test Split

This step prepares the engineered dataset for machine learning models. The goal is to convert categorical predictors into a format suitable for modeling and to divide the data into training and testing sets so that model performance can be evaluated on unseen data.

Key objectives of this stage:

	•	remove identifier and leakage columns
	•	encode categorical predictors
	•	split the dataset into training and testing sets
	•	preserve the class distribution of the target variable


## .Prepare Modeling Dataset

Identifier columns and raw outcome variables should not be used as predictors because they either uniquely identify records or directly contain the target outcome.

```{r}

# --------------------------------------------------
# Prepare Modeling Dataset
# --------------------------------------------------

model_dataset <- qal_features %>%
  select(
    -qal_id,
    -contact_lead_id,
    -next_stage_c,
    -contact_lead_title,
    -title_clean
  )

tibble(
  Metric = c("Records", "Columns"),
  Value = c(nrow(model_dataset), ncol(model_dataset))
) %>%
  gt() %>%
  tab_header(title = "Model Dataset Size")

```


## .Convert Target Variable to Factor

Although progressed is numeric, classification models typically expect a factor outcome.

```{r}
model_dataset <- model_dataset %>%
  mutate(
    progressed = factor(progressed, levels = c(0,1))
  )

```


## .Train/Test Split

The dataset is split into training and testing subsets. Stratified sampling ensures the proportion of progressed vs non-progressed leads remains consistent across both sets.

Typical split:

	•	70% training
	•	30% testing
	
	
```{r}
library(rsample)

set.seed(123)

data_split <- initial_split(
  model_dataset,
  prop = 0.7,
  strata = progressed
)

train_data <- training(data_split)
test_data  <- testing(data_split)

tibble(
  Dataset = c("Training Set", "Testing Set"),
  Records = c(nrow(train_data), nrow(test_data))
) %>%
  gt() %>%
  tab_header(title = "Train/Test Split Summary")
```


## .Verify Target Distribution After Split

This check confirms that stratification preserved the original class balance.

```{r}
train_data %>%
  count(progressed) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2)) %>%
  gt() %>%
  tab_header(title = "Training Set Target Distribution")

test_data %>%
  count(progressed) %>%
  mutate(Percentage = round(n / sum(n) * 100, 2)) %>%
  gt() %>%
  tab_header(title = "Testing Set Target Distribution")
```


## .Encode Categorical Predictors

Most predictors in this dataset are categorical variables. These must be encoded before modeling.

One-hot encoding converts each category into binary indicator variables.

```{r}
library(recipes)

# --------------------------------------------------
# Encoding Recipe
# --------------------------------------------------

encoding_recipe <- recipe(progressed ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_dummy(all_nominal_predictors())

encoding_prep <- prep(encoding_recipe)

train_encoded <- bake(encoding_prep, new_data = train_data)
test_encoded  <- bake(encoding_prep, new_data = test_data)

```



## .Final Modeling Dataset Check

```{r}
tibble(
  Dataset = c("Encoded Training Set", "Encoded Testing Set"),
  Records = c(nrow(train_encoded), nrow(test_encoded)),
  Columns = c(ncol(train_encoded), ncol(test_encoded))
) %>%
  gt() %>%
  tab_header(title = "Encoded Dataset Summary")

```




# . Modeling 


## . Descriptive Analysis of the Dataset 



	•	Cleaned and prepare the dataset, then analyze baseline conversion rates and segment-level performance for Mx and Qx across industry, manufacturing model, site function, company tier, territory, and buyer roles.




```{r}
# ==================================================
# Step 1 — Create Clean Analysis Datasets
# ==================================================

library(dplyr)
library(forcats)
library(tidyr)

# Helper: collapse rare levels
collapse_levels <- function(x, min_count = 50) {

  x <- as.character(x)

  tab <- table(x, useNA = "no")

  rare_levels <- names(tab[tab < min_count])

  x[x %in% rare_levels] <- "Other"

  factor(x)

}

analysis_train <- train_data %>%
  mutate(
    progressed = factor(progressed, levels = c(0, 1)),
    solution_rollup = factor(solution_rollup),
    acct_target_industry = collapse_levels(acct_target_industry),
    acct_manufacturing_model = collapse_levels(acct_manufacturing_model),
    acct_primary_site_function = collapse_levels(acct_primary_site_function),
    acct_tier_rollup = factor(acct_tier_rollup),
    acct_territory_rollup = collapse_levels(acct_territory_rollup),
    title_seniority = collapse_levels(title_seniority),
    title_function = collapse_levels(title_function)
  ) %>%
  select(
    progressed,
    solution_rollup,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  ) %>%
  droplevels()

analysis_test <- test_data %>%
  mutate(
    progressed = factor(progressed, levels = c(0, 1)),
    solution_rollup = factor(solution_rollup),
    acct_target_industry = collapse_levels(acct_target_industry),
    acct_manufacturing_model = collapse_levels(acct_manufacturing_model),
    acct_primary_site_function = collapse_levels(acct_primary_site_function),
    acct_tier_rollup = factor(acct_tier_rollup),
    acct_territory_rollup = collapse_levels(acct_territory_rollup),
    title_seniority = collapse_levels(title_seniority),
    title_function = collapse_levels(title_function)
  ) %>%
  select(
    progressed,
    solution_rollup,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  ) %>%
  droplevels()

cat("Rows in training analysis dataset:", nrow(analysis_train), "\n")
cat("Rows in testing analysis dataset:", nrow(analysis_test), "\n")
cat("Columns in analysis datasets:", ncol(analysis_train), "\n")


# ==================================================
# Step 2 — Descriptive Conversion Analysis
# ==================================================

library(gt)
library(tibble)

analysis_all <- bind_rows(analysis_train, analysis_test)

# Overall baseline
product_baseline <- analysis_all %>%
  filter(solution_rollup %in% c("Mx", "Qx")) %>%
  group_by(solution_rollup) %>%
  summarise(
    total_leads = n(),
    progressed_leads = sum(progressed == 1, na.rm = TRUE),
    conversion_rate = ifelse(total_leads > 0,
                             progressed_leads / total_leads,
                             NA),
    conversion_percent = round(conversion_rate * 100, 2),
    .groups = "drop"
  )

product_baseline %>%
  gt() %>%
  tab_header(title = "Overall Conversion Baseline — Mx vs Qx")

# Reusable function
segment_conversion <- function(data, segment_var) {
  data %>%
    filter(solution_rollup %in% c("Mx", "Qx")) %>%
    group_by(solution_rollup, .data[[segment_var]]) %>%
    summarise(
      total_leads = n(),
      progressed = sum(progressed == 1, na.rm = TRUE),
      conversion_rate = progressed / total_leads,
      conversion_percent = round(conversion_rate * 100, 2),
      .groups = "drop"
    ) %>%
    rename(Segment = .data[[segment_var]])
}

# Account-level views
segment_conversion(analysis_all, "acct_target_industry") %>%
  gt() %>%
  tab_header(title = "Conversion by Industry — Mx vs Qx")

segment_conversion(analysis_all, "acct_manufacturing_model") %>%
  gt() %>%
  tab_header(title = "Conversion by Manufacturing Model — Mx vs Qx")

segment_conversion(analysis_all, "acct_primary_site_function") %>%
  gt() %>%
  tab_header(title = "Conversion by Site Function — Mx vs Qx")

segment_conversion(analysis_all, "acct_tier_rollup") %>%
  gt() %>%
  tab_header(title = "Conversion by Company Tier — Mx vs Qx")

segment_conversion(analysis_all, "acct_territory_rollup") %>%
  gt() %>%
  tab_header(title = "Conversion by Territory — Mx vs Qx")

# Title-level views
segment_conversion(analysis_all, "title_seniority") %>%
  gt() %>%
  tab_header(title = "Conversion by Title Seniority — Mx vs Qx")

segment_conversion(analysis_all, "title_function") %>%
  gt() %>%
  tab_header(title = "Conversion by Title Function — Mx vs Qx")


# ==================================================
# Step 3 — Cross-Sell and Impact Analysis
# ==================================================

library(tidyr)

mx_baseline <- product_baseline %>%
  filter(solution_rollup == "Mx") %>%
  pull(conversion_rate)

qx_baseline <- product_baseline %>%
  filter(solution_rollup == "Qx") %>%
  pull(conversion_rate)

cross_sell_analysis <- function(data, segment_var) {
  data %>%
    filter(solution_rollup %in% c("Mx", "Qx")) %>%
    group_by(solution_rollup, .data[[segment_var]]) %>%
    summarise(
      total_leads = n(),
      progressed = sum(progressed == 1, na.rm = TRUE),
      conversion_rate = progressed / total_leads,
      .groups = "drop"
    ) %>%
    select(solution_rollup, .data[[segment_var]], conversion_rate) %>%
    pivot_wider(
      names_from = solution_rollup,
      values_from = conversion_rate
    ) %>%
    mutate(
      mx_conversion = round(Mx * 100, 2),
      qx_conversion = round(Qx * 100, 2),
      cross_sell_flag = ifelse(
        Mx > mx_baseline & Qx > qx_baseline,
        "High Cross-Sell Potential",
        "Low Cross-Sell Potential"
      )
    ) %>%
    rename(Segment = .data[[segment_var]])
}

cross_sell_analysis(analysis_all, "acct_manufacturing_model") %>%
  select(Segment, mx_conversion, qx_conversion, cross_sell_flag) %>%
  gt() %>%
  tab_header(title = "Cross-Sell Opportunity — Manufacturing Model")

cross_sell_analysis(analysis_all, "acct_primary_site_function") %>%
  select(Segment, mx_conversion, qx_conversion, cross_sell_flag) %>%
  gt() %>%
  tab_header(title = "Cross-Sell Opportunity — Site Function")

cross_sell_analysis(analysis_all, "title_seniority") %>%
  select(Segment, mx_conversion, qx_conversion, cross_sell_flag) %>%
  gt() %>%
  tab_header(title = "Cross-Sell Opportunity — Title Seniority")

cross_sell_analysis(analysis_all, "title_function") %>%
  select(Segment, mx_conversion, qx_conversion, cross_sell_flag) %>%
  gt() %>%
  tab_header(title = "Cross-Sell Opportunity — Title Function")

# Impact analysis
recommended_mx_segments <- analysis_all %>%
  filter(
    solution_rollup == "Mx",
    acct_tier_rollup %in% c("Small", "Medium"),
    acct_target_industry != "Non-Life Science",
    acct_primary_site_function != "Non-manufacturing organization"
  )

impact_table <- tibble(
  Metric = c(
    "Current Mx Baseline Conversion",
    "Recommended Segment Conversion",
    "Absolute Lift (percentage points)",
    "Relative Lift (%)"
  ),
  Value = c(
    round(mx_baseline * 100, 2),
    round(mean(recommended_mx_segments$progressed == 1) * 100, 2),
    round((mean(recommended_mx_segments$progressed == 1) - mx_baseline) * 100, 2),
    round(((mean(recommended_mx_segments$progressed == 1) - mx_baseline) / mx_baseline) * 100, 2)
  )
)
print(
impact_table %>%
  gt() %>%
  tab_header(title = "Estimated Impact of Improved Mx Targeting")
)
```





## . Model 1: Logistic Regression

This chunk builds an interpretable baseline model to estimate how each account and title characteristic affects the probability of progression, while also reporting coefficients, odds ratios, and classification performance.

  •	Build logistic regression models to predict the probability that a lead progresses (conversion = 1) based on account characteristics and title attributes.
	•	Evaluate model performance (AUC, precision, recall, F1) and identify the most influential predictors and high-performing segments to improve Mx targeting and cross-sell opportunities

```{r}
# ==================================================
# Logistic Modeling Function
# ==================================================

library(broom)
library(caret)
library(yardstick)
library(pROC)
library(pscl)
library(ggplot2)

run_logistic_model <- function(train_df, test_df, formula_obj, model_title) {

  target_var <- all.vars(formula_obj)[1]
  vars_needed <- all.vars(formula_obj)

  # Keep only rows complete for variables in the formula
  train_model_data <- train_df %>%
    select(all_of(vars_needed)) %>%
    drop_na()

  test_model_data <- test_df %>%
    select(all_of(vars_needed)) %>%
    drop_na()

  # Fit full and null models
  full_model <- glm(
    formula_obj,
    data = train_model_data,
    family = binomial
  )

  null_model <- glm(
    as.formula(paste(target_var, "~ 1")),
    data = train_model_data,
    family = binomial
  )

  # Likelihood ratio test
  lr_test <- anova(null_model, full_model, test = "Chisq")

  # Pseudo R2
  pseudo_r2 <- pscl::pR2(full_model)

  # Coefficients
  coef_tbl <- broom::tidy(full_model, conf.int = TRUE) %>%
    mutate(
      odds_ratio = exp(estimate),
      or_conf_low = exp(conf.low),
      or_conf_high = exp(conf.high),
      significance = case_when(
        p.value < 0.001 ~ "*** Highly Significant",
        p.value < 0.01  ~ "** Very Significant",
        p.value < 0.05  ~ "* Significant",
        TRUE ~ "Not Significant"
      )
    ) %>%
    filter(term != "(Intercept)") %>%
    arrange(p.value)

  # --------------------------------------------------
  # Align factor levels with model training levels
  # --------------------------------------------------

  for (v in names(full_model$xlevels)) {
    test_model_data[[v]] <- factor(
      test_model_data[[v]],
      levels = full_model$xlevels[[v]]
    )
  }

  # Remove rows with unseen levels
  test_model_data <- test_model_data %>%
    drop_na()

  cat("Rows remaining in test dataset after alignment:",
      nrow(test_model_data), "\n")

  # --------------------------------------------------
  # Predictions on test data
  # --------------------------------------------------

  test_model_data$pred_prob <- predict(
    full_model,
    newdata = test_model_data,
    type = "response"
  )

  # --------------------------------------------------
  # Calibration Plot
  # --------------------------------------------------

  calibration_df <- data.frame(
    prob = test_model_data$pred_prob,
    actual = as.numeric(as.character(test_model_data[[target_var]]))
  )

  print(
    ggplot(calibration_df, aes(x = prob, y = actual)) +
      geom_smooth(method = "loess", color = "#2C7FB8") +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      labs(
        title = paste("Calibration Plot —", model_title),
        x = "Predicted Probability",
        y = "Observed Outcome"
      ) +
      theme_minimal()
  )

  # --------------------------------------------------
  # Determine threshold using TRAIN data (avoid leakage)
  # --------------------------------------------------

  train_model_data$train_prob <- predict(
    full_model,
    newdata = train_model_data,
    type = "response"
  )

  roc_train <- roc(
    train_model_data[[target_var]],
    train_model_data$train_prob
  )

  optimal_threshold <- as.numeric(
    coords(roc_train, x = "best", ret = "threshold")
  )

  test_model_data$pred_class <- ifelse(
    test_model_data$pred_prob >= optimal_threshold,
    1,
    0
  )

  test_model_data$pred_class <- factor(test_model_data$pred_class, levels = c(0, 1))

  conf_matrix <- confusionMatrix(
    test_model_data$pred_class,
    test_model_data[[target_var]],
    positive = "1"
  )

  roc_obj <- roc(
    test_model_data[[target_var]],
    test_model_data$pred_prob
  )

  metrics_tbl <- tibble(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "AUC", "Threshold"),
    Value = c(
      accuracy_vec(test_model_data[[target_var]], test_model_data$pred_class),
      precision_vec(test_model_data[[target_var]], test_model_data$pred_class, event_level = "second"),
      recall_vec(test_model_data[[target_var]], test_model_data$pred_class, event_level = "second"),
      f_meas_vec(test_model_data[[target_var]], test_model_data$pred_class, event_level = "second"),
      as.numeric(auc(roc_obj)),
      optimal_threshold
    )
  )

  # Output
  cat("\n========================================\n")
  cat(model_title, "\n")
  cat("========================================\n\n")

  cat("Training rows used:", nrow(train_model_data), "\n")
  cat("Testing rows used:", nrow(test_model_data), "\n\n")

  cat("AIC:\n")
  print(AIC(full_model))

  cat("\nLikelihood Ratio Test:\n")
  print(lr_test)

  cat("\nPseudo R-squared:\n")
  print(pseudo_r2)

  print(
    coef_tbl %>%
      select(
        Predictor = term,
        Estimate = estimate,
        `Odds Ratio` = odds_ratio,
        `CI Lower` = or_conf_low,
        `CI Upper` = or_conf_high,
        `p-value` = p.value,
        Significance = significance
      ) %>%
      gt() %>%
      tab_header(title = paste(model_title, "— Significant Predictors")) %>%
      fmt_number(
        columns = c(Estimate, `Odds Ratio`, `CI Lower`, `CI Upper`, `p-value`),
        decimals = 4
      )
  )

  top_coef <- coef_tbl %>%
    filter(p.value < 0.10) %>%
    slice_head(n = 15)

  print(
    ggplot(top_coef, aes(x = reorder(term, odds_ratio), y = odds_ratio)) +
      geom_col(fill = "#2C7FB8") +
      coord_flip() +
      labs(
        title = paste(model_title, "— Top Predictors"),
        x = "Predictor",
        y = "Odds Ratio"
      ) +
      theme_minimal()
  )

  conf_df <- as.data.frame(conf_matrix$table)

  print(
    ggplot(conf_df, aes(x = Prediction, y = Reference, fill = Freq)) +
      geom_tile(color = "white", linewidth = 1) +
      geom_text(aes(label = Freq), size = 6) +
      scale_fill_gradient(low = "#DCE6F1", high = "#1F4E79") +
      labs(
        title = paste("Confusion Matrix —", model_title),
        x = "Predicted",
        y = "Actual"
      ) +
      theme_minimal()
  )

  print(
    metrics_tbl %>%
      gt() %>%
      tab_header(title = paste(model_title, "— Performance Metrics")) %>%
      fmt_number(columns = Value, decimals = 4)
  )

  plot(
    roc_obj,
    col = "#2C7FB8",
    lwd = 3,
    main = paste("ROC Curve —", model_title, "\nAUC =", round(as.numeric(auc(roc_obj)), 3))
  )
  abline(a = 0, b = 1, lty = 2)

  invisible(
    list(
      model = full_model,
      coefficients = coef_tbl,
      lr_test = lr_test,
      pseudo_r2 = pseudo_r2,
      confusion_matrix = conf_matrix,
      metrics = metrics_tbl,
      roc = roc_obj,
      threshold = optimal_threshold
    )
  )
}


# ==================================================
# Step 5 — Run Logistic Models for Deliverables
# ==================================================

set.seed(123)

# 5A. Account model for Mx
mx_account_train <- analysis_train %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup
  )

mx_account_test <- analysis_test %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup
  )

mx_account_results <- run_logistic_model(
  train_df = mx_account_train,
  test_df = mx_account_test,
  formula_obj = progressed ~ acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    acct_territory_rollup,
  model_title = "Mx Account Model"
)

# 5B. Title model for Mx
mx_title_train <- analysis_train %>%
  filter(solution_rollup == "Mx") %>%
  select(progressed, title_seniority, title_function)

mx_title_test <- analysis_test %>%
  filter(solution_rollup == "Mx") %>%
  select(progressed, title_seniority, title_function)

mx_title_results <- run_logistic_model(
  train_df = mx_title_train,
  test_df = mx_title_test,
  formula_obj = progressed ~ title_seniority + title_function,
  model_title = "Mx Title Model"
)

# 5C. Combined model for Mx
mx_combined_train <- analysis_train %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    title_seniority,
    title_function
  )

mx_combined_test <- analysis_test %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    title_seniority,
    title_function
  )

mx_combined_results <- run_logistic_model(
  train_df = mx_combined_train,
  test_df = mx_combined_test,
  formula_obj = progressed ~ acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    title_seniority +
    title_function,
  model_title = "Mx Combined Account + Title Model"
)

# 5D. Cross-product model: Mx vs Qx
cross_train <- analysis_train %>%
  filter(solution_rollup %in% c("Mx", "Qx")) %>%
  mutate(mx_product = factor(ifelse(solution_rollup == "Mx", 1, 0), levels = c(0,1))) %>%
  select(
    mx_product,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  )

cross_test <- analysis_test %>%
  filter(solution_rollup %in% c("Mx", "Qx")) %>%
  mutate(mx_product = factor(ifelse(solution_rollup == "Mx", 1, 0), levels = c(0,1))) %>%
  select(
    mx_product,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  )

cross_results <- run_logistic_model(
  train_df = cross_train,
  test_df = cross_test,
  formula_obj = mx_product ~ acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    acct_territory_rollup +
    title_seniority +
    title_function,
  model_title = "Mx vs Qx Cross-Sell Model"
)


```



## . Model 2: Elastic-Net Logistic Regression (glmnet)

This chunk extends logistic regression by adding regularization, which helps stabilize estimates, handle correlated predictors, and automatically shrink less useful effects in a feature-rich setting. 


```{r}
# ==================================================
# GLMNET Logistic Regression Function
# ==================================================

library(glmnet)
library(recipes)
library(dplyr)
library(caret)
library(yardstick)
library(pROC)
library(tibble)
library(gt)
library(ggplot2)
library(broom)

run_glmnet_model <- function(train_df,
                             test_df,
                             formula_obj,
                             model_title,
                             alpha_value = 1) {

  target_var <- all.vars(formula_obj)[1]
  vars_needed <- all.vars(formula_obj)

  # --------------------------------------------------
  # Keep only required variables
  # --------------------------------------------------

  train_model_data <- train_df %>%
    select(all_of(vars_needed)) %>%
    drop_na()

  test_model_data <- test_df %>%
    select(all_of(vars_needed)) %>%
    drop_na()

  train_model_data[[target_var]] <- factor(train_model_data[[target_var]], levels = c(0, 1))
  test_model_data[[target_var]]  <- factor(test_model_data[[target_var]], levels = c(0, 1))

  # --------------------------------------------------
  # Recipe for dummy encoding
  # --------------------------------------------------

  glmnet_recipe <- recipe(formula_obj, data = train_model_data) %>%
    step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
    step_zv(all_predictors())

  glmnet_prep <- prep(glmnet_recipe, training = train_model_data)

  train_baked <- bake(glmnet_prep, new_data = train_model_data)
  test_baked  <- bake(glmnet_prep, new_data = test_model_data)

  # --------------------------------------------------
  # Prepare x and y
  # --------------------------------------------------

  train_x <- train_baked %>%
    select(-all_of(target_var)) %>%
    as.matrix()

  test_x <- test_baked %>%
    select(-all_of(target_var)) %>%
    as.matrix()

  train_y <- as.numeric(as.character(train_baked[[target_var]]))
  test_y  <- as.numeric(as.character(test_baked[[target_var]]))

  # --------------------------------------------------
  # Cross-validated GLMNET
  # alpha = 1  -> LASSO
  # alpha = 0  -> Ridge
  # alpha = 0.5 -> Elastic Net
  # --------------------------------------------------

  cv_glmnet <- cv.glmnet(
    x = train_x,
    y = train_y,
    family = "binomial",
    alpha = alpha_value,
    nfolds = 5,
    type.measure = "auc"
  )

  best_lambda <- cv_glmnet$lambda.min

  glmnet_model <- glmnet(
    x = train_x,
    y = train_y,
    family = "binomial",
    alpha = alpha_value,
    lambda = best_lambda
  )

  # --------------------------------------------------
  # Predictions
  # --------------------------------------------------

  train_prob <- as.numeric(
    predict(glmnet_model, newx = train_x, type = "response")
  )

  test_prob <- as.numeric(
    predict(glmnet_model, newx = test_x, type = "response")
  )

  # --------------------------------------------------
  # Threshold from TRAIN ROC
  # --------------------------------------------------

  roc_train <- roc(train_y, train_prob)

  optimal_threshold <- as.numeric(
    coords(roc_train, x = "best", ret = "threshold")
  )

  test_pred_class <- ifelse(test_prob >= optimal_threshold, 1, 0)

  test_pred_class_factor <- factor(test_pred_class, levels = c(0, 1))
  test_y_factor <- factor(test_y, levels = c(0, 1))

  # --------------------------------------------------
  # Confusion Matrix
  # --------------------------------------------------

  conf_matrix <- confusionMatrix(
    test_pred_class_factor,
    test_y_factor,
    positive = "1"
  )

  # --------------------------------------------------
  # ROC + Metrics
  # --------------------------------------------------

  roc_obj <- roc(test_y, test_prob)

  metrics_tbl <- tibble(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "AUC", "Threshold", "Best Lambda"),
    Value = c(
      accuracy_vec(test_y_factor, test_pred_class_factor),
      precision_vec(test_y_factor, test_pred_class_factor, event_level = "second"),
      recall_vec(test_y_factor, test_pred_class_factor, event_level = "second"),
      f_meas_vec(test_y_factor, test_pred_class_factor, event_level = "second"),
      as.numeric(auc(roc_obj)),
      optimal_threshold,
      best_lambda
    )
  )

  # --------------------------------------------------
  # Coefficients
  # --------------------------------------------------

  coef_mat <- as.matrix(coef(glmnet_model))
  coef_df <- tibble(
    term = rownames(coef_mat),
    estimate = as.numeric(coef_mat[,1])
  ) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      odds_ratio = exp(estimate),
      abs_estimate = abs(estimate)
    ) %>%
    arrange(desc(abs_estimate))

  # --------------------------------------------------
  # Output
  # --------------------------------------------------

  cat("\n========================================\n")
  cat(model_title, "\n")
  cat("========================================\n\n")

  cat("Training rows used:", nrow(train_model_data), "\n")
  cat("Testing rows used:", nrow(test_model_data), "\n")
  cat("Alpha:", alpha_value, "\n")
  cat("Best Lambda:", best_lambda, "\n")
  cat("Threshold:", round(optimal_threshold, 4), "\n\n")

  print(
    metrics_tbl %>%
      gt() %>%
      tab_header(title = paste(model_title, "— Performance Metrics")) %>%
      fmt_number(columns = Value, decimals = 4)
  )

  print(
    coef_df %>%
      select(
        Predictor = term,
        Estimate = estimate,
        `Odds Ratio` = odds_ratio
      ) %>%
      slice_head(n = 20) %>%
      gt() %>%
      tab_header(title = paste(model_title, "— Top GLMNET Coefficients")) %>%
      fmt_number(columns = c(Estimate, `Odds Ratio`), decimals = 4)
  )

  top_coef <- coef_df %>%
    slice_head(n = 15)

  print(
    ggplot(top_coef, aes(x = reorder(term, abs_estimate), y = abs_estimate)) +
      geom_col(fill = "#2C7FB8") +
      coord_flip() +
      labs(
        title = paste(model_title, "— Most Important GLMNET Predictors"),
        x = "Predictor",
        y = "Absolute Coefficient"
      ) +
      theme_minimal()
  )

  conf_df <- as.data.frame(conf_matrix$table)

  print(
    ggplot(conf_df, aes(x = Prediction, y = Reference, fill = Freq)) +
      geom_tile(color = "white", linewidth = 1) +
      geom_text(aes(label = Freq), size = 6) +
      scale_fill_gradient(low = "#DCE6F1", high = "#1F4E79") +
      labs(
        title = paste("Confusion Matrix —", model_title),
        x = "Predicted",
        y = "Actual"
      ) +
      theme_minimal()
  )

  plot(
    roc_obj,
    col = "#2C7FB8",
    lwd = 3,
    main = paste("ROC Curve —", model_title, "\nAUC =", round(as.numeric(auc(roc_obj)), 3))
  )
  abline(a = 0, b = 1, lty = 2)

  invisible(
    list(
      model = glmnet_model,
      cv_model = cv_glmnet,
      coefficients = coef_df,
      confusion_matrix = conf_matrix,
      metrics = metrics_tbl,
      roc = roc_obj,
      threshold = optimal_threshold,
      lambda = best_lambda
    )
  )
}


# ==================================================
# Step 7 — Run GLMNET Models for Deliverables
# ==================================================

set.seed(123)

# 7A. GLMNET Account model for Mx
mx_account_glmnet <- run_glmnet_model(
  train_df = mx_account_train,
  test_df = mx_account_test,
  formula_obj = progressed ~ acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    acct_territory_rollup,
  model_title = "GLMNET — Mx Account Model",
  alpha_value = 1
)

# 7B. GLMNET Title model for Mx
mx_title_glmnet <- run_glmnet_model(
  train_df = mx_title_train,
  test_df = mx_title_test,
  formula_obj = progressed ~ title_seniority + title_function,
  model_title = "GLMNET — Mx Title Model",
  alpha_value = 1
)

# 7C. GLMNET Combined model for Mx
mx_combined_glmnet <- run_glmnet_model(
  train_df = mx_combined_train,
  test_df = mx_combined_test,
  formula_obj = progressed ~ acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    title_seniority +
    title_function,
  model_title = "GLMNET — Mx Combined Account + Title Model",
  alpha_value = 1
)

# 7D. GLMNET Cross-product model: Mx vs Qx
cross_glmnet <- run_glmnet_model(
  train_df = cross_train,
  test_df = cross_test,
  formula_obj = mx_product ~ acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    acct_territory_rollup +
    title_seniority +
    title_function,
  model_title = "GLMNET — Mx vs Qx Cross-Sell Model",
  alpha_value = 1
)
```




## . Model 3: Decision Tree (CART)


This chunk fits a rule-based model that splits the data into practical segments, making it easier to visualize how combinations of account and title factors lead to stronger or weaker progression.


```{r}
# ==================================================
# Model 3 — Decision Tree (CART)
# Drivers of Mx Lead Progression
# ==================================================

library(rpart)
library(rpart.plot)
library(dplyr)
library(ggplot2)
library(caret)
library(yardstick)
library(pROC)
library(gt)
library(tibble)

set.seed(123)

# --------------------------------------------------
# Prepare Training and Testing Data
# --------------------------------------------------

mx_tree_train <- analysis_train %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  ) %>%
  drop_na()

mx_tree_test <- analysis_test %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  ) %>%
  drop_na()

# Ensure target is factor
mx_tree_train$progressed <- factor(mx_tree_train$progressed, levels = c(0,1))
mx_tree_test$progressed  <- factor(mx_tree_test$progressed, levels = c(0,1))

# --------------------------------------------------
# Align Factor Levels Between Train/Test
# --------------------------------------------------

for(v in names(mx_tree_train)){
  
  if(is.factor(mx_tree_train[[v]])){
    
    mx_tree_test[[v]] <- factor(
      mx_tree_test[[v]],
      levels = levels(mx_tree_train[[v]])
    )
    
  }
}

mx_tree_test <- mx_tree_test %>% drop_na()

cat("Training rows:", nrow(mx_tree_train), "\n")
cat("Testing rows:", nrow(mx_tree_test), "\n")

# --------------------------------------------------
# Handle Class Imbalance
# --------------------------------------------------

class_weights <- ifelse(mx_tree_train$progressed == "1", 5, 1)

# --------------------------------------------------
# Train Decision Tree
# --------------------------------------------------

mx_tree_model <- rpart(
  
  progressed ~

    acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    acct_territory_rollup +
    title_seniority +
    title_function,

  data = mx_tree_train,
  method = "class",
  weights = class_weights,

  control = rpart.control(
    cp = 0.001,
    minsplit = 20,
    maxdepth = 6
  )
)

print(mx_tree_model)

# --------------------------------------------------
# Prune Tree using Cross-Validated Error
# --------------------------------------------------

best_cp <- mx_tree_model$cptable[
  which.min(mx_tree_model$cptable[,"xerror"]),
  "CP"
]

mx_tree_pruned <- rpart::prune(mx_tree_model, cp = best_cp)

# --------------------------------------------------
# Visualize Decision Tree
# --------------------------------------------------

rpart.plot(
  mx_tree_pruned,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  tweak = 1.3,
  box.palette = "GnBu",
  main = "Decision Tree — Drivers of Mx Lead Progression"
)

# --------------------------------------------------
# Generate Predictions
# --------------------------------------------------

train_prob <- predict(
  mx_tree_pruned,
  newdata = mx_tree_train,
  type = "prob"
)[,2]

mx_tree_test$pred_prob <- predict(
  mx_tree_pruned,
  newdata = mx_tree_test,
  type = "prob"
)[,2]

# --------------------------------------------------
# Determine Optimal Threshold (TRAIN DATA)
# --------------------------------------------------

roc_train <- roc(mx_tree_train$progressed, train_prob)

optimal_threshold <- as.numeric(
  coords(roc_train, x = "best", ret = "threshold")
)

cat("Optimal Threshold:", optimal_threshold, "\n")

# --------------------------------------------------
# Apply Threshold to Test Predictions
# --------------------------------------------------

mx_tree_test$pred_class <- ifelse(
  mx_tree_test$pred_prob >= optimal_threshold,
  1,
  0
)

mx_tree_test$pred_class <- factor(mx_tree_test$pred_class, levels = c(0,1))

# --------------------------------------------------
# Confusion Matrix
# --------------------------------------------------

conf_matrix_tree <- confusionMatrix(
  mx_tree_test$pred_class,
  mx_tree_test$progressed,
  positive = "1"
)

conf_matrix_tree

# --------------------------------------------------
# Model Performance Metrics
# --------------------------------------------------

metrics_tree <- tibble(
  Metric = c("Accuracy","Precision","Recall","F1 Score"),
  Value = c(
    accuracy_vec(mx_tree_test$progressed, mx_tree_test$pred_class),
    precision_vec(mx_tree_test$progressed, mx_tree_test$pred_class, event_level="second"),
    recall_vec(mx_tree_test$progressed, mx_tree_test$pred_class, event_level="second"),
    f_meas_vec(mx_tree_test$progressed, mx_tree_test$pred_class, event_level="second")
  )
)

metrics_tree %>%
  gt() %>%
  tab_header(title = "Decision Tree Model Performance Metrics") %>%
  fmt_number(columns = Value, decimals = 4)

# --------------------------------------------------
# ROC Curve
# --------------------------------------------------

roc_tree <- roc(mx_tree_test$progressed, mx_tree_test$pred_prob)
auc_tree <- auc(roc_tree)

plot(
  roc_tree,
  col = "#2C7FB8",
  lwd = 3,
  main = paste("ROC Curve — Decision Tree\nAUC =", round(auc_tree,3))
)

abline(a = 0, b = 1, lty = 2)

# --------------------------------------------------
# Confusion Matrix Visualization
# --------------------------------------------------

conf_df <- as.data.frame(conf_matrix_tree$table)

ggplot(conf_df,
       aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 6) +
  scale_fill_gradient(low = "#DCE6F1", high = "#1F4E79") +
  labs(
    title = "Confusion Matrix — Decision Tree",
    x = "Predicted",
    y = "Actual"
  ) +
  theme_minimal()

# --------------------------------------------------
# Variable Importance
# --------------------------------------------------

var_imp <- tibble(
  Variable = names(mx_tree_pruned$variable.importance),
  Importance = as.numeric(mx_tree_pruned$variable.importance)
)

var_imp %>%
  arrange(desc(Importance)) %>%
  gt() %>%
  tab_header(title = "Decision Tree Variable Importance")

ggplot(var_imp,
       aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Variable Importance — Decision Tree",
    x = "Predictor",
    y = "Importance"
  ) +
  theme_minimal()
```





## . Model 4: Random Forest


This chunk builds many trees and averages them to improve robustness, reduce overfitting, and identify the most important predictors driving Mx progression.


```{r}
# ==================================================
# Model 4 — Random Forest
# Drivers of Mx Lead Progression
# ==================================================

library(randomForest)
library(dplyr)
library(caret)
library(pROC)
library(yardstick)
library(ggplot2)
library(gt)
library(tibble)

set.seed(123)

# --------------------------------------------------
# Prepare Training and Test Data
# --------------------------------------------------

mx_rf_train <- analysis_train %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  ) %>%
  drop_na()

mx_rf_test <- analysis_test %>%
  filter(solution_rollup == "Mx") %>%
  select(
    progressed,
    acct_target_industry,
    acct_manufacturing_model,
    acct_primary_site_function,
    acct_tier_rollup,
    acct_territory_rollup,
    title_seniority,
    title_function
  ) %>%
  drop_na()

# ensure target is factor
mx_rf_train$progressed <- factor(mx_rf_train$progressed, levels = c(0,1))
mx_rf_test$progressed  <- factor(mx_rf_test$progressed, levels = c(0,1))

# --------------------------------------------------
# Align Factor Levels Between Train/Test
# --------------------------------------------------

for(v in names(mx_rf_train)){

  if(is.factor(mx_rf_train[[v]])){

    mx_rf_test[[v]] <- factor(
      mx_rf_test[[v]],
      levels = levels(mx_rf_train[[v]])
    )

  }

}

mx_rf_test <- mx_rf_test %>% drop_na()

cat("Training rows:", nrow(mx_rf_train), "\n")
cat("Testing rows:", nrow(mx_rf_test), "\n")

# --------------------------------------------------
# Train Random Forest
# --------------------------------------------------

mx_rf_model <- randomForest(

  progressed ~

    acct_target_industry +
    acct_manufacturing_model +
    acct_primary_site_function +
    acct_tier_rollup +
    acct_territory_rollup +
    title_seniority +
    title_function,

  data = mx_rf_train,

  ntree = 500,
  mtry = 3,
  importance = TRUE,

  # handle class imbalance
  classwt = c("0" = 1, "1" = 4)

)

print(mx_rf_model)

# --------------------------------------------------
# Predictions (Train + Test)
# --------------------------------------------------

train_prob <- predict(
  mx_rf_model,
  newdata = mx_rf_train,
  type = "prob"
)[,2]

mx_rf_test$pred_prob <- predict(
  mx_rf_model,
  newdata = mx_rf_test,
  type = "prob"
)[,2]

# --------------------------------------------------
# ROC and AUC
# --------------------------------------------------

roc_rf <- roc(mx_rf_test$progressed, mx_rf_test$pred_prob)
auc_rf <- auc(roc_rf)

plot(
  roc_rf,
  col="#1F78B4",
  lwd=3,
  main=paste("ROC Curve — Random Forest\nAUC =", round(auc_rf,3))
)

abline(a=0,b=1,lty=2)

# --------------------------------------------------
# Optimal Threshold (FROM TRAIN DATA)
# --------------------------------------------------

roc_train <- roc(mx_rf_train$progressed, train_prob)

optimal_threshold <- as.numeric(
  coords(roc_train, x="best", ret="threshold")
)

cat("Optimal Threshold:", optimal_threshold, "\n")

# --------------------------------------------------
# Apply Threshold
# --------------------------------------------------

mx_rf_test$pred_class <- ifelse(
  mx_rf_test$pred_prob >= optimal_threshold,
  1,
  0
)

mx_rf_test$pred_class <- factor(mx_rf_test$pred_class, levels=c(0,1))

# --------------------------------------------------
# Confusion Matrix
# --------------------------------------------------

conf_matrix_rf <- confusionMatrix(
  mx_rf_test$pred_class,
  mx_rf_test$progressed,
  positive="1"
)

conf_matrix_rf

# --------------------------------------------------
# Performance Metrics
# --------------------------------------------------

metrics_rf <- tibble(
  Metric=c("Accuracy","Precision","Recall","F1 Score","AUC"),
  Value=c(
    accuracy_vec(mx_rf_test$progressed, mx_rf_test$pred_class),
    precision_vec(mx_rf_test$progressed, mx_rf_test$pred_class, event_level="second"),
    recall_vec(mx_rf_test$progressed, mx_rf_test$pred_class, event_level="second"),
    f_meas_vec(mx_rf_test$progressed, mx_rf_test$pred_class, event_level="second"),
    as.numeric(auc_rf)
  )
)

metrics_rf %>%
  gt() %>%
  tab_header(title="Random Forest Model Performance") %>%
  fmt_number(columns=Value, decimals=4)

# --------------------------------------------------
# Variable Importance
# --------------------------------------------------

var_imp <- importance(mx_rf_model)

var_imp_df <- tibble(
  Variable = rownames(var_imp),
  MeanDecreaseAccuracy = var_imp[,"MeanDecreaseAccuracy"],
  MeanDecreaseGini = var_imp[,"MeanDecreaseGini"]
)

var_imp_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  gt() %>%
  tab_header(title="Random Forest Variable Importance")

ggplot(var_imp_df,
       aes(x=reorder(Variable, MeanDecreaseAccuracy),
           y=MeanDecreaseAccuracy)) +
  geom_col(fill="#1F78B4") +
  coord_flip() +
  labs(
    title="Variable Importance — Random Forest",
    x="Predictor",
    y="Importance"
  ) +
  theme_minimal()
```




## . Model 5: Gradient Boosting / XGBoost

This chunk fits a boosted tree model that focuses on predictive strength and captures nonlinear relationships and interactions that simpler models may miss.  


```{r}
# ==================================================
# Model 5 — XGBoost
# Feature-Engineered Predictive Model
# ==================================================

library(dplyr)
library(tidyr)
library(recipes)
library(xgboost)
library(pROC)
library(caret)
library(yardstick)
library(tibble)
library(gt)
library(ggplot2)
library(scales)

set.seed(123)

# --------------------------------------------------
# Reusable XGBoost Function
# --------------------------------------------------

run_xgb_model <- function(train_df,
                          test_df,
                          outcome_var,
                          predictor_vars,
                          model_title,
                          nfold = 5,
                          nrounds = 500,
                          early_stopping_rounds = 30,
                          eta = 0.05,
                          max_depth = 5,
                          min_child_weight = 8,
                          subsample = 0.8,
                          colsample_bytree = 0.8,
                          gamma = 1) {

  # -----------------------------------------------
  # Keep only required columns
  # -----------------------------------------------

  train_model_data <- train_df %>%
    select(all_of(c(outcome_var, predictor_vars))) %>%
    drop_na()

  test_model_data <- test_df %>%
    select(all_of(c(outcome_var, predictor_vars))) %>%
    drop_na()

  # make outcome numeric 0/1 factor
  train_model_data[[outcome_var]] <- factor(train_model_data[[outcome_var]], levels = c(0, 1))
  test_model_data[[outcome_var]]  <- factor(test_model_data[[outcome_var]], levels = c(0, 1))

  # -----------------------------------------------
  # Recipe for encoding engineered predictors
  # -----------------------------------------------

  xgb_recipe <- recipe(
    as.formula(paste(outcome_var, "~", paste(predictor_vars, collapse = " + "))),
    data = train_model_data
  ) %>%
    step_other(all_nominal_predictors(), threshold = 0.03) %>%
    step_unknown(all_nominal_predictors()) %>%
    step_novel(all_nominal_predictors()) %>%
    step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
    step_zv(all_predictors())

  xgb_prep <- prep(xgb_recipe, training = train_model_data)

  train_encoded <- bake(xgb_prep, new_data = train_model_data)
  test_encoded  <- bake(xgb_prep, new_data = test_model_data)

  # -----------------------------------------------
  # Create x and y
  # -----------------------------------------------

  train_y <- as.numeric(as.character(train_encoded[[outcome_var]]))
  test_y  <- as.numeric(as.character(test_encoded[[outcome_var]]))

  train_x <- train_encoded %>%
    select(-all_of(outcome_var)) %>%
    as.matrix()

  test_x <- test_encoded %>%
    select(-all_of(outcome_var)) %>%
    as.matrix()

  dtrain <- xgb.DMatrix(data = train_x, label = train_y)
  dtest  <- xgb.DMatrix(data = test_x, label = test_y)

  # -----------------------------------------------
  # Class imbalance handling
  # -----------------------------------------------

  neg_count <- sum(train_y == 0)
  pos_count <- sum(train_y == 1)
  scale_pos_weight_value <- neg_count / pos_count

  # -----------------------------------------------
  # Parameters
  # -----------------------------------------------

  params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = eta,
    max_depth = max_depth,
    min_child_weight = min_child_weight,
    subsample = subsample,
    colsample_bytree = colsample_bytree,
    gamma = gamma,
    scale_pos_weight = scale_pos_weight_value
  )

  # -----------------------------------------------
  # Cross-validation
  # -----------------------------------------------

  cv_model <- xgb.cv(
    params = params,
    data = dtrain,
    nrounds = nrounds,
    nfold = nfold,
    stratified = TRUE,
    early_stopping_rounds = early_stopping_rounds,
    verbose = 0,
    maximize = TRUE
  )

  best_nrounds <- max(cv_model$best_iteration, 1)

  # -----------------------------------------------
  # Final model
  # -----------------------------------------------

  xgb_model <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = best_nrounds,
    verbose = 0
  )

  # -----------------------------------------------
  # Predictions
  # -----------------------------------------------

  train_pred_prob <- predict(xgb_model, dtrain)
  test_pred_prob  <- predict(xgb_model, dtest)

  # -----------------------------------------------
  # Threshold from TRAIN ROC
  # -----------------------------------------------

  train_roc <- roc(response = train_y, predictor = train_pred_prob, quiet = TRUE)

  optimal_threshold <- as.numeric(
    coords(train_roc, x = "best", ret = "threshold")
  )

  test_pred_class <- ifelse(test_pred_prob >= optimal_threshold, 1, 0)

  test_pred_class_factor <- factor(test_pred_class, levels = c(0, 1))
  test_y_factor <- factor(test_y, levels = c(0, 1))

  # -----------------------------------------------
  # Confusion matrix
  # -----------------------------------------------

  conf_matrix <- confusionMatrix(
    test_pred_class_factor,
    test_y_factor,
    positive = "1"
  )

  # -----------------------------------------------
  # Metrics
  # -----------------------------------------------

  test_roc <- roc(response = test_y, predictor = test_pred_prob, quiet = TRUE)
  auc_value <- as.numeric(auc(test_roc))

  metrics_tbl <- tibble(
    Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "AUC", "Threshold", "Best CV Rounds"),
    Value = c(
      accuracy_vec(test_y_factor, test_pred_class_factor),
      precision_vec(test_y_factor, test_pred_class_factor, event_level = "second"),
      recall_vec(test_y_factor, test_pred_class_factor, event_level = "second"),
      f_meas_vec(test_y_factor, test_pred_class_factor, event_level = "second"),
      auc_value,
      optimal_threshold,
      best_nrounds
    )
  )

  # -----------------------------------------------
  # Feature importance
  # -----------------------------------------------

  importance_tbl <- xgb.importance(
    feature_names = colnames(train_x),
    model = xgb_model
  )

  # -----------------------------------------------
  # Threshold tuning table on TEST set
  # -----------------------------------------------

  threshold_grid <- seq(0.10, 0.70, by = 0.05)

  threshold_tbl <- lapply(threshold_grid, function(th) {
    pred_class <- factor(ifelse(test_pred_prob >= th, 1, 0), levels = c(0, 1))

    tibble(
      Threshold = th,
      Accuracy = accuracy_vec(test_y_factor, pred_class),
      Precision = precision_vec(test_y_factor, pred_class, event_level = "second"),
      Recall = recall_vec(test_y_factor, pred_class, event_level = "second"),
      F1 = f_meas_vec(test_y_factor, pred_class, event_level = "second")
    )
  }) %>%
    bind_rows()

  # -----------------------------------------------
  # Output section
  # -----------------------------------------------

  cat("\n==================================================\n")
  cat(model_title, "\n")
  cat("==================================================\n\n")

  cat("Training rows used:", nrow(train_model_data), "\n")
  cat("Testing rows used :", nrow(test_model_data), "\n")
  cat("Scale pos weight  :", round(scale_pos_weight_value, 3), "\n")
  cat("Best CV rounds    :", best_nrounds, "\n")
  cat("Train-based threshold:", round(optimal_threshold, 4), "\n\n")

  print(
    metrics_tbl %>%
      gt() %>%
      tab_header(title = paste(model_title, "— Performance Metrics")) %>%
      fmt_number(columns = Value, decimals = 4)
  )

  conf_df <- as.data.frame(conf_matrix$table)

  print(
    ggplot(conf_df, aes(x = Prediction, y = Reference, fill = Freq)) +
      geom_tile(color = "white", linewidth = 1) +
      geom_text(aes(label = Freq), size = 6) +
      scale_fill_gradient(low = "#DCE6F1", high = "#1F4E79") +
      labs(
        title = paste("Confusion Matrix —", model_title),
        x = "Predicted",
        y = "Actual"
      ) +
      theme_minimal()
  )

  plot(
    test_roc,
    col = "#2C7FB8",
    lwd = 3,
    main = paste("ROC Curve —", model_title, "\nAUC =", round(auc_value, 3))
  )
  abline(a = 0, b = 1, lty = 2)

  print(
    importance_tbl %>%
      slice_head(n = 15) %>%
      ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
      geom_col(fill = "#2C7FB8") +
      coord_flip() +
      labs(
        title = paste(model_title, "— Top Features by Gain"),
        x = "Feature",
        y = "Gain"
      ) +
      theme_minimal()
  )

  # threshold tuning plot
  threshold_long <- threshold_tbl %>%
    pivot_longer(cols = c(Accuracy, Precision, Recall, F1),
                 names_to = "Metric",
                 values_to = "Value")

  print(
    ggplot(threshold_long, aes(x = Threshold, y = Value, color = Metric)) +
      geom_line(linewidth = 1) +
      geom_point(size = 2) +
      labs(
        title = paste("Threshold Optimization —", model_title),
        x = "Threshold",
        y = "Metric Value"
      ) +
      theme_minimal()
  )

  print(
    threshold_tbl %>%
      gt() %>%
      tab_header(title = paste(model_title, "— Threshold Tuning Table")) %>%
      fmt_number(columns = c(Accuracy, Precision, Recall, F1), decimals = 4)
  )

  invisible(list(
    model = xgb_model,
    metrics = metrics_tbl,
    confusion_matrix = conf_matrix,
    roc = test_roc,
    threshold_table = threshold_tbl,
    importance = importance_tbl
  ))
}

# ==================================================
# Step 1 — XGBoost Mx Account Model
# ==================================================

mx_account_train <- train_data %>%
  filter(solution_rollup == "Mx")

mx_account_test <- test_data %>%
  filter(solution_rollup == "Mx")

mx_account_xgb <- run_xgb_model(
  train_df = mx_account_train,
  test_df = mx_account_test,
  outcome_var = "progressed",
  predictor_vars = c(
    "acct_target_industry",
    "acct_manufacturing_model",
    "acct_primary_site_function",
    "acct_tier_rollup",
    "acct_territory_rollup"
  ),
  model_title = "XGBoost — Mx Account Model"
)

# ==================================================
# Step 2 — XGBoost Mx Title Model
# ==================================================

mx_title_xgb <- run_xgb_model(
  train_df = mx_account_train,
  test_df = mx_account_test,
  outcome_var = "progressed",
  predictor_vars = c(
    "title_seniority",
    "title_function"
  ),
  model_title = "XGBoost — Mx Title Model"
)

# ==================================================
# Step 3 — XGBoost Mx Combined Model
# ==================================================

mx_combined_xgb <- run_xgb_model(
  train_df = mx_account_train,
  test_df = mx_account_test,
  outcome_var = "progressed",
  predictor_vars = c(
    "acct_target_industry",
    "acct_manufacturing_model",
    "acct_primary_site_function",
    "acct_tier_rollup",
    "acct_territory_rollup",
    "title_seniority",
    "title_function"
  ),
  model_title = "XGBoost — Mx Combined Account + Title Model"
)

# ==================================================
# Step 4 — XGBoost Mx vs Qx Cross-Sell Model
# ==================================================

cross_train <- train_data %>%
  filter(solution_rollup %in% c("Mx", "Qx")) %>%
  mutate(mx_product = ifelse(solution_rollup == "Mx", 1, 0))

cross_test <- test_data %>%
  filter(solution_rollup %in% c("Mx", "Qx")) %>%
  mutate(mx_product = ifelse(solution_rollup == "Mx", 1, 0))

cross_xgb <- run_xgb_model(
  train_df = cross_train,
  test_df = cross_test,
  outcome_var = "mx_product",
  predictor_vars = c(
    "acct_target_industry",
    "acct_manufacturing_model",
    "acct_primary_site_function",
    "acct_tier_rollup",
    "acct_territory_rollup",
    "title_seniority",
    "title_function"
  ),
  model_title = "XGBoost — Mx vs Qx Cross-Sell Model"
)
```



# . Model Comparison

This chunk compares models using classification metrics such as accuracy, precision, recall, F1, and ROC-AUC so the final recommendation balances performance with interpretability.


```{r}
# ==================================================
# Model Comparison — Performance Summary
# ==================================================

library(dplyr)
library(tidyr)
library(gt)

# --------------------------------------------------
# Extract metrics from each model
# --------------------------------------------------

logistic_metrics <- mx_combined_results$metrics %>%
  mutate(Model = "Logistic Regression")

glmnet_metrics <- mx_combined_glmnet$metrics %>%
  mutate(Model = "Elastic Net (GLMNET)")

tree_metrics <- metrics_tree %>%
  mutate(Model = "Decision Tree")

rf_metrics <- metrics_rf %>%
  mutate(Model = "Random Forest")

xgb_metrics <- mx_combined_xgb$metrics %>%
  mutate(Model = "XGBoost")

# --------------------------------------------------
# Combine all model metrics
# --------------------------------------------------

model_comparison <- bind_rows(
  logistic_metrics,
  glmnet_metrics,
  tree_metrics,
  rf_metrics,
  xgb_metrics
)

# --------------------------------------------------
# Pivot to comparison table
# --------------------------------------------------

comparison_table <- model_comparison %>%
  select(Model, Metric, Value) %>%
  pivot_wider(
    names_from = Metric,
    values_from = Value
  )

# --------------------------------------------------
# Display comparison table
# --------------------------------------------------

comparison_table %>%
  gt() %>%
  tab_header(
    title = "Model Performance Comparison",
    subtitle = "Evaluation using Accuracy, Precision, Recall, F1 Score, and ROC-AUC"
  ) %>%
  fmt_number(
    columns = where(is.numeric),
    decimals = 4
  )


# --------------------------------------------------
# Visualization of Model Performance
# --------------------------------------------------

comparison_long <- model_comparison %>%
  filter(Metric %in% c("Accuracy","Precision","Recall","F1 Score","AUC"))

library(ggplot2)

ggplot(comparison_long,
       aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "Model Performance Comparison",
    x = "Model",
    y = "Metric Value"
  ) +
  theme_minimal()



```






# . Executive Analysis and Final Recommendations

## . Overview

The objective of this analysis was to understand why the Mx product line progresses qualified leads at a lower rate than Qx, and to develop a data-driven targeting framework that improves Mx conversion performance. The analysis combined descriptive analytics, predictive modeling, and prescriptive recommendations to translate historical lead progression data into actionable commercial strategy.

Across the analysis, several consistent patterns emerged. First, the performance gap between Mx and Qx is not primarily explained by product capability but by targeting misalignment. Mx is inherently tied to manufacturing execution and operational workflows, meaning its adoption depends heavily on whether an organization actually operates manufacturing processes and whether the relevant decision-makers are being targeted. Second, the data shows that manufacturing context and operational buyer roles are stronger indicators of progression than broad industry labels or geographic segmentation.

The modeling phase evaluated multiple predictive approaches, including logistic regression, elastic-net logistic regression, decision trees, random forest, and XGBoost. While more complex models offered slightly stronger predictive flexibility, Logistic regression therefore provides a clear link between statistical results and business decision-making, allowing MasterControl to identify which account and buyer characteristics increase or decrease the probability of Mx progression, with elastic-net logistic regression used as a supporting validation model.

The findings from this analysis provide a clear framework for improving Mx targeting and strengthening overall commercial effectiveness.


## . Descriptive Analytics: Understanding Current Performance Patterns

The descriptive phase examined how lead progression varies across several business dimensions, including product line, industry, manufacturing model, site function, company tier, geography, and buyer titles.

### . Product Performance Gap

The first key observation is the performance difference between the two products. Qx demonstrates consistently stronger progression rates, while Mx shows lower conversion across most segments. This confirms that the issue is not isolated to a single industry or territory but reflects a broader structural pattern.

### . Importance of Manufacturing Context

Manufacturing model emerged as the strongest account-level driver of progression. Organizations operating in-house manufacturing environments or contract manufacturing models consistently showed stronger performance compared with organizations that primarily operate as service providers, consulting organizations, or non-manufacturing entities.

This result highlights an important structural issue in the current targeting strategy. Because Mx is a manufacturing execution solution, outreach directed toward organizations without meaningful manufacturing operations is unlikely to produce strong progression results.

### . Role of Buyer Personas

Buyer characteristics also influence progression outcomes. Titles associated with Director and Manager levels often demonstrated stronger progression than generic executive outreach. Additionally, functional roles tied to quality management, manufacturing operations, and operational oversight performed more consistently than roles unrelated to manufacturing processes.

These patterns suggest that the most effective Mx buyers are often operational leaders responsible for execution and workflow management rather than purely strategic executive stakeholders.

### . Industry and Geography

Industry segmentation alone does not fully explain progression outcomes. While life sciences industries remain the natural focus for MasterControl, the analysis indicates that manufacturing structure and buyer relevance are more important predictors than industry classification alone.

Similarly, geographic segmentation does not appear to meaningfully differentiate progression performance. This suggests that improving targeting precision should focus more on operational structure and persona alignment rather than geographic prioritization.



## . Predictive Analytics: Identifying the Drivers of Mx Progression

The predictive phase translated the descriptive patterns into statistical models designed to estimate the probability that a lead progresses through the sales pipeline.

Several models were evaluated:

	-	Logistic Regression
	-	Elastic-Net Logistic Regression (GLMNET)
	-	Decision Tree (CART)
	-	Random Forest
	-	Gradient Boosting (XGBoost)

Each model provides different advantages. Tree-based and boosting models can capture complex interactions and nonlinear relationships, while logistic models provide clearer interpretability.

Why Logistic Regression Was Selected as the Primary Model

Although all models produced useful insights, logistic regression was selected as the primary model for several reasons:
	1.	Interpretability
Logistic regression clearly quantifies how each account or title attribute influences progression probability. This makes it possible to translate statistical outputs into business decisions.
	2.	Alignment with Client Deliverables
The client’s key questions require interpretable relationships between account characteristics and lead progression. Logistic regression allows the analysis to directly answer questions such as:
	-	which industries convert best,
	-	which manufacturing models drive progression,
	-	which job titles represent the strongest buyers.
	3.	Transparency for Commercial Teams
Sales and marketing teams must understand why certain accounts or personas should be prioritized. Logistic regression provides a transparent explanation framework, making it easier to operationalize the results.

Role of Elastic-Net Logistic Regression

Elastic-net logistic regression was used as a supporting model to validate the findings from the primary logistic regression model. Because elastic-net introduces regularization, it helps stabilize estimates and reduce the impact of correlated predictors.

The elastic-net results largely confirmed the same variables identified by logistic regression, strengthening confidence that the observed drivers of progression are not simply artifacts of model specification.

Role of Tree-Based Models

Decision trees, random forests, and XGBoost models were valuable for benchmarking predictive performance and identifying nonlinear patterns. However, they are less interpretable and therefore less suitable as the primary decision framework for this project. Instead, they serve as supporting models that validate the core patterns discovered through logistic analysis.


**Table: Role of Each Model in the Analysis**

| Model | Purpose in Analysis |
|:---|:---|
| Logistic Regression | Primary interpretable model used to identify key drivers of lead progression |
| Elastic Net (GLMNET) | Validation model that applies regularization to stabilize estimates and handle correlated predictors |
| Decision Tree (CART) | Rule-based segmentation model used to identify interpretable decision patterns |
| Random Forest | Ensemble model used to evaluate variable importance and reduce overfitting |
| XGBoost | High-performance gradient boosting model used as a predictive benchmark |


## . Prescriptive Analytics: Translating Findings into Business Strategy

The final stage of the analysis translates the descriptive and predictive findings into concrete business recommendations. This section directly addresses the client’s requested deliverables.


# . Client Deliverables and Strategic Recommendations

## .Deliverable 1: Which Account Types Should MasterControl Target for Mx?

MasterControl offers two primary enterprise software solutions within its platform: Qx (Quality Excellence) and Mx (Manufacturing Excellence). Qx digitizes quality and compliance processes such as document control, training management, audits, CAPA, and regulatory workflows. Because regulatory compliance is mandatory across many life sciences organizations, Qx has broad applicability across both manufacturing and non-manufacturing companies. 

Mx, in contrast, is a manufacturing execution solution (MES) that digitally manages production workflows, electronic batch records, and embedded quality checks within manufacturing processes. Unlike Qx, the value of Mx depends heavily on whether the organization actually operates manufacturing environments. As a result, the potential market for Mx is structurally narrower than that of Qx.  

The descriptive and predictive analysis consistently indicates that the strongest opportunities for Mx occur in organizations where manufacturing activity is central to operations.

Based on the analysis, MasterControl should prioritize:

- organizations operating in-house manufacturing environments
- contract manufacturing organizations (CMOs/CDMOs)
- companies where manufacturing execution and production workflows are core operational activities

These organizations have clear operational needs for digital production management, batch traceability, and manufacturing process control, which aligns directly with the value proposition of Mx.

In contrast, accounts that primarily operate as service organizations, consulting firms, or non-manufacturing entities should generally be deprioritized for Mx outreach unless there is explicit evidence that they operate manufacturing facilities.

Industry targeting should remain focused on regulated life sciences sectors, particularly pharmaceutical, biotechnology, and medical device organizations. However, the analysis shows that manufacturing structure is a stronger predictor of Mx progression than industry classification alone.

Company size also influences progression patterns. Small and medium-tier organizations often demonstrate strong potential because they tend to have fewer legacy systems and may adopt digital manufacturing solutions more quickly than very large enterprises.

Overall, the data suggests that manufacturing relevance is the primary account-level indicator of Mx opportunity, far more important than broad industry or geographic segmentation.



## .Deliverable 2: Which Job Titles Convert Best for Mx?

While account structure determines whether an organization is a good candidate for Mx, buyer personas determine whether a lead progresses through the sales pipeline.

The analysis indicates that the most effective Mx buyers are mid-to-senior operational leaders responsible for manufacturing and production processes.

The strongest converting personas include:

- Directors of Manufacturing
- Directors of Quality
- Manufacturing Operations Managers
- Quality Systems Managers
- Operational Excellence Leaders

These roles are closely tied to production workflows, manufacturing performance, and operational compliance. As a result, they are more likely to recognize the operational efficiency and traceability benefits provided by Mx.

This finding aligns with the fundamental positioning of Mx as a manufacturing execution platform. Because the product embeds quality checks directly into manufacturing processes and digitizes batch execution workflows, the individuals responsible for day-to-day operations are the most natural buyers.

Senior executive roles, such as C-level leaders, may still influence final purchasing decisions, but the analysis suggests that initial progression is more strongly associated with operational decision-makers who directly experience manufacturing challenges.

Therefore, MasterControl should prioritize operational leaders early in the sales process, while engaging executive stakeholders later in the buying cycle.



## .Deliverable 3: How Does the Ideal Mx Buyer Differ from the Ideal Qx Buyer?

The analysis reveals a fundamental difference between the buyer profiles of Qx and Mx.

Qx serves as a quality management system (QMS) that supports compliance activities such as document control, training, audit readiness, and risk management. Because regulatory compliance requirements apply broadly across regulated industries, Qx can be adopted by a wide range of organizations regardless of whether they actively manufacture products.  ￼

As a result, Qx buyers tend to come from quality assurance, regulatory, and compliance functions, and the solution converts well across both manufacturing and non-manufacturing organizations.

Mx, on the other hand, supports manufacturing execution and production traceability, which makes it highly dependent on operational manufacturing environments. Its buyers are therefore more specialized.

Key differences between the two buyer profiles include:

**Table: Comparison of Qx and Mx Buyer Profiles**

| Dimension | Qx Buyer Profile | Mx Buyer Profile |
|:---|:---|:---|
| Core Value | Compliance & audit readiness | Digital manufacturing execution |
| Manufacturing Requirement | Not required | Required |
| Primary Buyer Functions | Quality / Regulatory | Manufacturing / Operations |
| Sensitivity to Site Function | Low | Very High |



Because of these differences, applying the same targeting strategy to both products can dilute Mx performance. Qx can succeed with broad compliance-oriented outreach, while Mx requires more focused targeting toward operational manufacturing environments.

## . Deliverable 4: Which Accounts and Titles Represent Cross-Sell Opportunities?

Cross-sell opportunities occur when both product value propositions align within the same organization.

The strongest cross-sell candidates are organizations that already maintain formal quality systems while also operating active manufacturing environments. In these cases, Because many regulated manufacturers first adopt quality management solutions before digitizing manufacturing execution, Qx deployments can naturally serve as an entry point for future Mx adoption.

The best cross-sell scenarios include:

- organizations already adopting quality management systems but operating manufacturing facilities
- operational leaders responsible for both manufacturing performance and quality compliance
- companies transitioning from quality-focused digital systems toward fully integrated manufacturing execution platforms

In these cases, Qx adoption may naturally create demand for Mx capabilities, particularly when organizations seek to link quality management with production workflows.

This cross-product relationship reflects the architecture of the MasterControl platform, which connects quality management, manufacturing execution, and asset management within a unified compliance-focused system.

The cross-sell model confirmed that manufacturing-active accounts already engaging with quality management workflows represent the strongest overlap segment for simultaneous Mx and Qx adoption.

## .Deliverable 5: What Is the Expected Impact of Improved Targeting?

By improving targeting precision, MasterControl can significantly improve the efficiency of Mx sales efforts.

Current performance suggests that a substantial portion of Mx leads originate from accounts or personas that are not strongly aligned with the product’s core use case. When outreach is directed toward non-manufacturing organizations or purely quality-focused roles, the probability of progression declines.

The model identified 310 high-priority Mx leads representing the accounts and personas most aligned with the Mx use case. This Tier 1 segment converted at 32.9% versus the 13.5% baseline — a +19.4 percentage point lift. In practical terms, MasterControl's SDRs should focus first on manufacturing-active accounts at small-to-mid tier company sizes, targeting operational leaders in manufacturing and quality functions, rather than applying broad outreach across all qualified leads. Even a partial reallocation of sales effort toward these Tier 1 accounts can produce meaningful pipeline gains without requiring additional lead generation volume.


## . Business Value of the Analytical Framework

Beyond answering the immediate business questions, this analysis establishes a repeatable framework for commercial decision-making.

The models and segmentation insights developed in this project can support:

- improved lead scoring models
- better segmentation of outbound marketing campaigns
- clearer definitions of Ideal Customer Profiles (ICPs) for both Qx and Mx
- stronger alignment between sales, marketing, and product teams

Most importantly, this framework shifts MasterControl’s approach from volume-based lead generation to precision targeting based on empirical progression data.

Instead of focusing primarily on generating more leads, the company can focus on identifying the right organizations, the right operational environments, and the right buyer personas for Mx adoption.


# . Mx Lead Prioritization Framework

This section translates the predictive modeling results into a direct action framework for MasterControl's sales team. Using the combined logistic regression model, all Mx leads are scored and ranked into three priority tiers based on their estimated probability of progression.

```{r}
# Setup and scoring
library(dplyr)
library(gt)

mx_all_leads <- analysis_all %>%
  filter(solution_rollup == "Mx") %>%
  select(progressed, acct_target_industry, acct_manufacturing_model,
    acct_primary_site_function, acct_tier_rollup, acct_territory_rollup,
    title_seniority, title_function) %>%
  drop_na()

for(v in names(mx_combined_results$model$xlevels)) {
  mx_all_leads[[v]] <- factor(mx_all_leads[[v]],
    levels = mx_combined_results$model$xlevels[[v]])
}

mx_all_leads <- mx_all_leads %>% drop_na()

mx_all_leads$pred_prob <- predict(mx_combined_results$model,
  newdata = mx_all_leads, type = "response")

mx_all_leads <- mx_all_leads %>%
  mutate(priority_tier = case_when(
    pred_prob >= 0.25 ~ "Tier 1 — High Priority",
    pred_prob >= 0.15 ~ "Tier 2 — Medium Priority",
    TRUE             ~ "Tier 3 — Low Priority"))
```

```{r}
tier_summary <- mx_all_leads %>%
  group_by(priority_tier) %>%
  summarise(Lead_Count = n(),
    Actual_Conversion = round(mean(progressed == 1) * 100, 1),
    .groups = "drop") %>%
  mutate(
    Pct_of_Pipeline  = round(Lead_Count / sum(Lead_Count) * 100, 1),
    Lift_vs_Baseline = round(Actual_Conversion - mean(mx_all_leads$progressed == 1) * 100, 1))

tier_summary %>%
  gt() %>%
  tab_header(title = "Mx Lead Prioritization Framework",
    subtitle = "Model-driven tiers showing expected conversion lift vs. baseline")
```

```{r}
tier1_profile <- mx_all_leads %>%
  filter(priority_tier == "Tier 1 — High Priority") %>%
  summarise(
    Top_Industry  = names(sort(table(acct_target_industry),    decreasing=TRUE))[1],
    Top_Mfg_Model = names(sort(table(acct_manufacturing_model),decreasing=TRUE))[1],
    Top_Site_Func = names(sort(table(acct_primary_site_function),decreasing=TRUE))[1],
    Top_Tier      = names(sort(table(acct_tier_rollup),        decreasing=TRUE))[1],
    Top_Seniority = names(sort(table(title_seniority),         decreasing=TRUE))[1],
    Top_Function  = names(sort(table(title_function),          decreasing=TRUE))[1])

tier1_profile %>%
  gt() %>%
  tab_header(title = "Tier 1 Lead Profile — Most Common Characteristics")
```

```{r}
baseline_conv <- round(mean(mx_all_leads$progressed == 1) * 100, 1)
tier1_conv    <- tier_summary %>%
  filter(priority_tier == "Tier 1 — High Priority") %>% pull(Actual_Conversion)
tier1_count   <- tier_summary %>%
  filter(priority_tier == "Tier 1 — High Priority") %>% pull(Lead_Count)

tibble(
  Metric = c("Overall Mx Baseline Conversion", "Tier 1 Conversion Rate",
    "Lift (percentage points)", "Tier 1 Lead Count"),
  Value  = c(paste0(baseline_conv, "%"), paste0(tier1_conv, "%"),
    paste0("+", tier1_conv - baseline_conv, " pp"), as.character(tier1_count))
) %>%
  gt() %>%
  tab_header(title = "Estimated Impact of Tier 1 Targeting",
    subtitle = "What MasterControl should prioritize differently going forward")
```


# . Final Conclusion

The analysis confirms that Mx underperformance is largely driven by targeting misalignment rather than inherent product limitations.

Organizations with active manufacturing operations and operational decision-makers represent the strongest opportunities for Mx adoption. By prioritizing these accounts and personas, MasterControl can improve conversion efficiency and better align commercial effort with actual product-market fit.

Among the predictive models evaluated, logistic regression and elastic-net logistic regression provide the best balance between predictive performance and interpretability. Logistic regression serves as the primary analytical model because it clearly explains the relationship between account characteristics, buyer roles, and progression outcomes, making it the most useful framework for guiding business decisions.

In practical terms, the analysis demonstrates that improving account fit and buyer alignment can significantly strengthen Mx pipeline efficiency without requiring additional lead generation, allowing MasterControl to scale growth through smarter targeting rather than larger lead volumes.

By applying these insights, MasterControl can improve the precision of its Mx targeting strategy, increase lead progression efficiency, and strengthen long-term adoption of the Mx product line.



# . Group Member Contributions


This project was completed through collaborative work by all group members. Each member independently developed a modeling file exploring different analytical approaches to address the business problem. These individual modeling efforts included experimentation with multiple predictive techniques and interpretation of results related to lead progression for the Mx and Qx product lines.

Following this process, the group reviewed the outputs collectively and selected the modeling approach that best aligned with the project objectives and the client deliverables. The final analysis presented in this report integrates insights from these individual modeling efforts and reflects the collective evaluation of model interpretability, business relevance, and analytical rigor.

In addition to modeling, group members contributed to data preparation, exploratory analysis, interpretation of results, development of business recommendations, and preparation of the final report.



# . AI Log

Artificial Intelligence tools were used in a limited and supportive capacity during this project.

AI assistance was primarily used to help with:

	•	debugging and refining R code,
	•	structuring certain visualization components,
	•	verifying interpretation of statistical outputs, and
	•	improving clarity in documentation and explanation of analytical results.

All modeling decisions, interpretation of results, and final recommendations were reviewed and validated by the project team to ensure accuracy and alignment with the analytical objectives of the project.

